---
title: "3 Cluster analysis"
author: "jeff karnsomprot"
date: "2025-07-28"
output: html_document
---


# Load Packages

```{r load new packages, echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}

if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("Seurat")) {install.packages("Seurat"); require("Seurat")}
if (!require("patchwork")) {install.packages("patchwork"); require("patchwork")}
if (!require("knitr")) {install.packages("knitr"); require("knitr")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
if (!require("BiocManager")) {install.packages("BiocManager"); require("BiocManager")}
if (!require("tibble")) {install.packages("tibble"); require("tibble")}
if (!require("ggpmisc")) {install.packages("ggpmisc"); require("ggpmisc")}
if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require("RColorBrewer")} #color
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("DESeq2")) {BiocManager::install('DESeq2'); require("DESeq2")}
if (!require("here")) {install.packages("here"); require("here")}
if (!require("stringr")) {install.packages("stringr"); require("stringr")}
if (!require("car")) {install.packages("car"); require("car")}
if (!require("openxlsx")) {install.packages("openxlsx"); require("openxlsx")}
if (!require("readxl")) {install.packages("readxl"); require("readxl")}
if (!require("data.table")) {install.packages("data.table"); require("data.table")}
if (!require("ggvenn")) {install.packages("ggvenn"); require("ggvenn")}
if (!require("kableExtra")) {install.packages("kableExtra"); require("kableExtra")} # for color brewer
if (!require("gplots")) {install.packages("gplots"); require("gplots")} # for color brewer
if (!require("clusterProfiler")) {BiocManager::install('clusterProfiler'); require("clusterProfiler")}
if (!require("enrichplot")) {BiocManager::install('enrichplot'); require("enrichplot")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")} # for data frame transformation
if (!require("plotly")) {install.packages("plotly"); require("plotly")}
library("EnhancedVolcano")

# Install from scratch
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("RSQLite")  # Core dependency
BiocManager::install("org.Mm.eg.db")
BiocManager::install("clusterProfiler")


library(org.Mm.eg.db)
library(clusterProfiler)
library(AnnotationDbi)
library(dplyr)
library(magrittr)




library("EnhancedVolcano")
library(UpSetR)


```

```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
options(future.globals.maxSize = 74 * 1024^3) # 55 GB
getOption("future.globals.maxSize") #59055800320
```

## Load Dataset

```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}

load(here("jk_code", "macula_densa_final.rds"))


```


# If 3 clusters?

```{r}

ElbowPlot(seu)
seu <- SCTransform(seu) %>%
    RunPCA() %>%
    FindNeighbors(dims = 1:12) %>%
    FindClusters(resolution = 1) %>%
    RunUMAP(dims = 1:12)

DimPlot(seu)



```
# Find Markers

```{r}


all_markers <- FindAllMarkers(seu, only.pos = TRUE, min.pct = 0.5, logfc.threshold = 0.5)

all_markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 0.5, p_val_adj < 0.05) %>%
    slice_head(n = 3) %>%
    ungroup() -> top3

DoHeatmap(seu, features = top3$gene) + NoLegend()

```
```{r}



seu@meta.data <- seu@meta.data %>%
  mutate(
    md_typev2 = dplyr::case_when(
      seurat_clusters %in% c(0,1,2,3,4,5,6,11,12,13) ~ "type_1",
      seurat_clusters %in% c(7,8,10,15) ~ "type_2",
      seurat_clusters %in% c(9,14) ~ "type_3"
  )
  )

DimPlot(seu, group.by = "md_typev2")



Idents(seu) <- "md_typev2"

type_markers <- FindAllMarkers(seu, 
                          only.pos = TRUE,
                          logfc.threshold = 0.1,
                          min.pct = 0.1,
                          min.diff.pct = 0.05,
                          return.thresh = 0.05)

type_markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1, p_val_adj < 0.05) %>%
    slice_head(n = 3) %>%
    ungroup() -> top3type

DoHeatmap(seu, features = top3type$gene) + NoLegend()

```


## Creating Pathway Function

```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}

#  Filter Before Independently before you Run this

# Creating a function to Repeat Easily

pathway_analysis <- function(type_markers, type_name)
  {
  # Arrange markers by log2 fold change
  df2 <- type_markers %>% arrange(desc(avg_log2FC))
  DEG_list <- df2
  
  markers <- DEG_list %>% mutate(SYMBOL = gene)


ENTREZ_list <- bitr(
  geneID = DEG_list$gene,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Mm.eg.db
)

  markers <- ENTREZ_list %>% inner_join(markers, by = "SYMBOL")
  
  # Upregulated Pathway Code
  pos_markers_up <- markers %>% dplyr::filter(avg_log2FC > 0) %>% arrange(desc(abs(avg_log2FC)))
  pos_ranks_up <- pos_markers_up$ENTREZID[abs(pos_markers_up$avg_log2FC) > 0]
  pos_go_up <- enrichGO(gene = pos_ranks_up, OrgDb = "org.Mm.eg.db", ont = "BP", readable = TRUE)
  chart_up <- dotplot(pos_go_up) +
    ggtitle(paste(type_name, "Functions")) +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5),
      legend.position = "left",
      axis.text.y = element_text(hjust = 0, size = 10)
    ) +
    scale_y_discrete(position = "right", labels = function(x) str_wrap(x, width = 25))
  

  return(list(
    chart_up = chart_up,
    pos_go_up = pos_go_up
  ))
}

```

## Pathways for Each MD Type

```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
# Creating List of Marker genes to do Pathway Analysis
type1_markers <- type_markers[type_markers$cluster == "type_1", ]
type2_markers <- type_markers[type_markers$cluster == "type_2", ]
type3_markers <- type_markers[type_markers$cluster == "type_3", ]


# Creating a Pos Go and Chart using PA function
results_type1 <- pathway_analysis(subset(type1_markers, avg_log2FC > 0.5), "type 1")
results_type2 <- pathway_analysis(subset(type2_markers, avg_log2FC > 0.85), "type 2")
results_type3 <- pathway_analysis(subset(type3_markers, avg_log2FC > 0), "type 3")


```

#### Type 1

```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
results_type1
```

#### Type 2

```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
results_type2
```

#### Type 3

```{r echo=TRUE, error=FALSE, message=FALSE, warning=FALSE}
results_type3
```


I checked a few cluster-specific genes from our existing MD-specific database that are very interesting, for example:

Fos, Socs3, Gadd45b, Wee1, Hspa1a, Sat1 for type 3

Cxcl12, Itpr2, Bmp4, Casr, Grin2c, Irx3, Rap1gap, App, Wwc1 for type 2

Syt5, Syn3, Cacna1d, Slc6a7, Robo2, Begain for type 1

   

Other genes like Pappa1, Nos1, Ptgs2, Bmp2, Atp2a3 may identify further subclusters within type 1.


```{r}

VlnPlot(seu,"Slc6a7",group.by = "md_type")

```