---
title: "macula densaan analysis"
author: "jeff karnsomprot"
date: "2025-04-18"
output:
  html_document: 
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: yes
    theme: journal
    df_print: paged
    code_folding: hide
    highlight: pygments

---
# Macula Densa Project

## Objectives 

Three Main Goals of this File <br>
Produce Cleaner looking code. <br> 
Identify the amount of clusters there are <br>
Identify the top genes expressed in each of the clusters <br>


## Problems I need to Fix
Save things as RDS file so I dont have to rerun the whole code

# Loading in Data sets + Library packages.

```{r load new packages, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require("knitr")) {install.packages("knitr"); require("knitr")}
if (!requireNamespace('BiocManager', quietly = TRUE)) {install.packages('BiocManager'); require("BiocManager")}
if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("stringr")) {install.packages("stringr"); require("stringr")}
if (!require("Seurat")) {install.packages("Seurat"); require("Seurat")}
if (!require("sctransform")) {install.packages("sctransform"); require("sctransform")}
if (!require("glmGamPoi")) {BiocManager::install('glmGamPoi'); require("glmGamPoi")}
if (!require("patchwork")) {install.packages("patchwork"); require("patchwork")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
if (!require("EnhancedVolcano")) {BiocManager::install('EnhancedVolcano'); require("EnhancedVolcano")}
if (!require("DESeq2")) {BiocManager::install('DESeq2'); require("DESeq2")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")}
if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require("RColorBrewer")}
if (!require("car")) {install.packages("car"); require("car")}
if (!require("openxlsx")) {install.packages("openxlsx"); require("openxlsx")}
if (!require("readxl")) {install.packages("readxl"); require("readxl")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("gghighlight")) {install.packages("gghighlight"); require("gghighlight")}
if (!require("ggpmisc")) {install.packages("ggpmisc"); require("ggpmisc")}
if (!require("data.table")) {install.packages("data.table"); require("data.table")}
if (!require("here")) {install.packages("here"); require("here")}
if (!require("clusterProfiler")) {BiocManager::install('clusterProfiler'); require("clusterProfiler")}
if (!require("enrichplot")) {BiocManager::install('enrichplot'); require("enrichplot")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")} # for data frame transformation
if (!require("gplots")) {install.packages("gplots"); require("gplots")}
if (!require("ggvenn")) {install.packages("ggvenn"); require("ggvenn")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}


# Remove corrupted installation
remove.packages(c("RSQLite", "org.Mm.eg.db", "clusterProfiler"))

# Install from scratch
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("RSQLite")  # Core dependency
BiocManager::install("org.Mm.eg.db")
BiocManager::install("clusterProfiler")


# CRAN packages
install.packages(c("dplyr", "magrittr"))
library(org.Mm.eg.db)
library(clusterProfiler)
library(AnnotationDbi)
library(dplyr)
library(magrittr)




library("EnhancedVolcano")
library(UpSetR)

```


```{r}
options(future.globals.maxSize = 74 * 1024^3) # 55 GB
getOption("future.globals.maxSize") #59055800320
```

``` {r}
load(here("jk_code", "JK_cleanMD.rds"))

```

# Analyzing DATASET
``` {r}

DimPlot(SO4,split.by ="sample",group.by = "sample")
DimPlot(SO4)

ElbowPlot(SO4)
SO4 <- SCTransform(SO4) %>%
    RunPCA() %>%
    FindNeighbors(dims = 1:15) %>%
    FindClusters(resolution = 0.5) %>%
    RunUMAP(dims = 1:15)

DimPlot(SO4)

```

## Viewing for different DEGs

```{r}
markers.to.plot_MD <- c(
  "Egf",
  "S100g",
  "Fos",
  "Foxq1",
  "Pappa2",
  "Mcub",
  "Hmox1")


DotPlot(SO4,
features = markers.to.plot_MD,
dot.scale = 8,
dot.min = 0,
scale = FALSE,
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5)+
coord_flip()


markers.to.plot1 <- c(
  "S100g",    # 
  "Atf3",     # 
  "Egr1",     # 
  "Fos",      # 
  "Jun",      #
  "Junb",     #
  "Pappa2",   #
  "Cxcl10",   # 
  "Cldn19",   # 
  "Krt7",     #
  "Egf",      #
  "Aard",
  "Ptger3",
  "Leng9",
  "Ckb",
  "Mcub",
  "Fabp3",
  "Ccn1",
  "Foxq1",
  "Cxcl12",
  "Vash2",
  "Pamr1",
  "Vegfa"
)


                      
                      
DotPlot(SO4,
features = markers.to.plot1,
dot.scale = 8,
dot.min = 0,
scale = FALSE,
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5)+
coord_flip()

DimPlot(SO4)

```

## Renaming my clusters 

```{r}


SO4@meta.data <- SO4@meta.data %>% 
  mutate(subclass_MD = dplyr::case_when(
    seurat_clusters == 0  ~ "type_1",
    seurat_clusters == 1  ~ "type_1",
    seurat_clusters == 2  ~ "type_1",
    seurat_clusters == 3  ~ "type_2",
    seurat_clusters == 4  ~ "type_1",
    seurat_clusters == 5  ~ "type_3",
    seurat_clusters == 6  ~ "type_4",
    seurat_clusters == 7  ~ "type_2"
  ))

SO4@meta.data$subclass_MD <- factor(SO4@meta.data$subclass_MD , levels = c("type_1", "type_2", "type_3","type_4","type_5"))

Idents(SO4) <- SO4@meta.data$subclass_MD

DimPlot(object = SO4, reduction = "umap", group.by = "subclass_MD", label = TRUE)


DimPlot(object = SO4, reduction = "umap", label = TRUE)

Idents(SO4) <- "subclass_MD"

DimPlot(SO4,group.by = "seurat_clusters")
DimPlot(SO4,group.by = "subclass_MD")

```
```{r}

markers.to.plot3<- c(
  "Pappa2",
  "Egf",
  "Jun",
  "S100g"
)


                      
                      
DotPlot(SO4,
features = markers.to.plot3,
dot.scale = 8,
dot.min = 0,
scale = FALSE,
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5)+
coord_flip()

```


## Top genes 


```{r}

# Find markers for all clusters (adjust parameters as needed)

SO.markers <- FindAllMarkers(SO4, only.pos = TRUE)
SO.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)

SO.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10
DoHeatmap(SO4, features = top10$gene) + NoLegend()

type_1_markers <- SO.markers[SO.markers$cluster == "type_1", ]
type_2_markers <- SO.markers[SO.markers$cluster == "type_2", ]
type_3_markers <- SO.markers[SO.markers$cluster == "type_3", ]
type_4_markers <- SO.markers[SO.markers$cluster == "type_4", ]

```

# Save Object
```{r}
save(SO4, file = here("jk_code", "SO4analysis.rds"))

```