---
title: "macula densa subtype analysis "
author: "jeff karnsomprot"
date: "2025-04-18"
output:
  html_document: 
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: yes
    theme: journal
    df_print: paged
    code_folding: hide
    highlight: pygments

---
# Macula Densa Project

## Objectives 

Three Main Goals of this File <br>
Produce Cleaner looking code. <br> 
Identify the amount of clusters there are <br>
Identify the top genes expressed in each of the clusters <br>


# Loading in Data sets + Library packages.

```{r load new packages, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require("knitr")) {install.packages("knitr"); require("knitr")}
if (!requireNamespace('BiocManager', quietly = TRUE)) {install.packages('BiocManager'); require("BiocManager")}
if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("stringr")) {install.packages("stringr"); require("stringr")}
if (!require("Seurat")) {install.packages("Seurat"); require("Seurat")}
if (!require("sctransform")) {install.packages("sctransform"); require("sctransform")}
if (!require("glmGamPoi")) {BiocManager::install('glmGamPoi'); require("glmGamPoi")}
if (!require("patchwork")) {install.packages("patchwork"); require("patchwork")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
if (!require("EnhancedVolcano")) {BiocManager::install('EnhancedVolcano'); require("EnhancedVolcano")}
if (!require("DESeq2")) {BiocManager::install('DESeq2'); require("DESeq2")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")}
if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require("RColorBrewer")}
if (!require("car")) {install.packages("car"); require("car")}
if (!require("openxlsx")) {install.packages("openxlsx"); require("openxlsx")}
if (!require("readxl")) {install.packages("readxl"); require("readxl")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("gghighlight")) {install.packages("gghighlight"); require("gghighlight")}
if (!require("ggpmisc")) {install.packages("ggpmisc"); require("ggpmisc")}
if (!require("data.table")) {install.packages("data.table"); require("data.table")}
if (!require("here")) {install.packages("here"); require("here")}
if (!require("clusterProfiler")) {BiocManager::install('clusterProfiler'); require("clusterProfiler")}
if (!require("enrichplot")) {BiocManager::install('enrichplot'); require("enrichplot")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")} # for data frame transformation
if (!require("gplots")) {install.packages("gplots"); require("gplots")}
if (!require("ggvenn")) {install.packages("ggvenn"); require("ggvenn")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}

# Install from scratch
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("RSQLite")  # Core dependency
BiocManager::install("org.Mm.eg.db")
BiocManager::install("clusterProfiler")


# CRAN packages
install.packages(c("dplyr", "magrittr"))
library(org.Mm.eg.db)
library(clusterProfiler)
library(AnnotationDbi)
library(dplyr)
library(magrittr)




library("EnhancedVolcano")
library(UpSetR)

```


```{r}
options(future.globals.maxSize = 74 * 1024^3) # 55 GB
getOption("future.globals.maxSize") #59055800320
```


``` {r}
load(here("jk_code", "JK_cleanMD_NEW.rds"))

```

# Analyzing DATASET

``` {r}


SO4@meta.data$sample <- factor(
  SO4@meta.data$sample, 
  levels = c("SO1","SO4","SO2","SO3"))
  
  
DimPlot(SO4,split.by ="sample",group.by = "sample")
DimPlot(SO4,group.by = "subclass2_MD",split.by = "sample")
DimPlot(SO4,group.by = "subclass_MD",split.by = "treatment")
DimPlot(SO4,group.by = "subclass2_MD",split.by = "treatment")


DimPlot(SO4)

```

## Viewing for different DEGs

```{r}


markers.to.plot1 <- c(
  "Atf3",     # 
  "Egr1",     # 
  "Fos",      # 
  "Jun",      #
  "Junb",     #
  "Pappa2",   #
  "Cxcl10",   # 
  "Cldn19",   # 
  "Krt7",     #
  "Egf",      #
  "Ptger3",
  "Ckb",
  "Mcub",
  "Fabp3",
  "Foxq1",
  "Vash2",
  "Pamr1",
  "Vegfa"
)


                      
DotPlot(SO4,
features = markers.to.plot1,
dot.scale = 8,
dot.min = 0,
scale = FALSE,
scale.max = 100,
scale.min = 0,
col.min = -2.5,
group.by = "seurat_clusters",
col.max = 2.5)+
coord_flip()

DimPlot(SO4,group.by = "seurat_clusters")
DimPlot(SO4,split.by = "treatment",group.by = "seurat_clusters")


library(plotly)

dim_plot <- DimPlot(SO4,split.by = "treatment",group.by = "seurat_clusters") +
  ggtitle("UMAP Dimensional Reduction")
interactive_plot <- ggplotly(dim_plot)

# Display the interactive plot
interactive_plot

```

## Finding Markers for each MD Subtype

```{r}

Idents(SO4) <- "subclass2_MD"

MD_DEGs <- FindAllMarkers(SO4, 
                          only.pos = TRUE,
                          logfc.threshold = 0.1,
                          min.pct = 0.1,
                          min.diff.pct = 0.1,
                          return.thresh = 0.05)
 
MD_DEGs <- MD_DEGs %>% arrange(desc(avg_log2FC))

write.xlsx(MD_DEGs, file = here("jk_code", "DEGs_MD_SUBclass.xlsx"), rowNames = TRUE)

# save as excel file

MD_DEGs %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1, p_val_adj < 0.05) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10


# save as excel file

DoHeatmap(SO4, features = top10$gene) + NoLegend()

top10

```

##  Markers for each MD Sub-SUb type 

```{r}

Idents(SO4) <- "subclass_MD"

MD_DEGs2 <- FindAllMarkers(SO4, 
                          only.pos = TRUE,
                          logfc.threshold = 0.1,
                          min.pct = 0.1,
                          min.diff.pct = 0.1,
                          return.thresh = 0.05)

MD_DEGs2 <- MD_DEGs2 %>% arrange(desc(avg_log2FC))

write.xlsx(MD_DEGs2, file = here("jk_code", "DEGs_MD_SUBSUBclass.xlsx"), rowNames = TRUE)

# save as excel file

MD_DEGs2 %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1, p_val_adj < 0.05) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10


# save as excel file

DoHeatmap(SO4, features = top10$gene) + NoLegend()


```

## Finding markers between each type now

### Type 1a Markers

```{r}

type_1 <- c("Pappa2","Ramp3","Itga4","Aard")

VlnPlot(SO4,type_1,)


```

### Type 2a Markers

```{r}

type_2a <- c("Egf","Umod","Cxcl12","Ckb")

VlnPlot(SO4,type_2a)


```

### Type 2b Markers

```{r}

type_2b <- c("Foxq1","Slc9a3","Sult1d1","Tmem52b")

VlnPlot(SO4,type_2b)


```

### Type 3 Markers

```{r}

type_3 <- c("Jun","Fos","Egr1","Hspa1a")

VlnPlot(SO4,type_3)


```


### Type 4 Markers

```{r}

type_4 <- c("Cxcl10","Isg15","Ifit1")

VlnPlot(SO4,type_4)


```


## Top Genes for each type 

```{r}

type_markers <- c(type_1, type_2a, type_2b, type_3, type_4)

                      
                      
DotPlot(SO4,
features = type_markers,
dot.scale = 8,
dot.min = 0,
scale = FALSE,
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5)+
coord_flip()

```


# Save Object
```{r}
save(SO4, file = here("jk_code", "SO4_SUB_NEW_analysis.rds"))

```