---
title: "upset_plot_wip"
author: "jeff karnsomprot"
date: "2025-07-08"
output:
  html_document: 
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: yes
    theme: journal
    df_print: paged
    code_folding: hide
    highlight: pygments
---

# Loading in Data sets + Library packages.

```{r load new packages, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require("knitr")) {install.packages("knitr"); require("knitr")}
if (!requireNamespace('BiocManager', quietly = TRUE)) {install.packages('BiocManager'); require("BiocManager")}
if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("stringr")) {install.packages("stringr"); require("stringr")}
if (!require("Seurat")) {install.packages("Seurat"); require("Seurat")}
if (!require("sctransform")) {install.packages("sctransform"); require("sctransform")}
if (!require("glmGamPoi")) {BiocManager::install('glmGamPoi'); require("glmGamPoi")}
if (!require("patchwork")) {install.packages("patchwork"); require("patchwork")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
if (!require("EnhancedVolcano")) {BiocManager::install('EnhancedVolcano'); require("EnhancedVolcano")}
if (!require("DESeq2")) {BiocManager::install('DESeq2'); require("DESeq2")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")}
if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require("RColorBrewer")}
if (!require("car")) {install.packages("car"); require("car")}
if (!require("openxlsx")) {install.packages("openxlsx"); require("openxlsx")}
if (!require("readxl")) {install.packages("readxl"); require("readxl")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("gghighlight")) {install.packages("gghighlight"); require("gghighlight")}
if (!require("ggpmisc")) {install.packages("ggpmisc"); require("ggpmisc")}
if (!require("data.table")) {install.packages("data.table"); require("data.table")}
if (!require("here")) {install.packages("here"); require("here")}
if (!require("clusterProfiler")) {BiocManager::install('clusterProfiler'); require("clusterProfiler")}
if (!require("enrichplot")) {BiocManager::install('enrichplot'); require("enrichplot")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")} # for data frame transformation
if (!require("gplots")) {install.packages("gplots"); require("gplots")}
if (!require("ggvenn")) {install.packages("ggvenn"); require("ggvenn")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("UpSetR")) {install.packages("UpSetR"); require("UpSetR")}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ComplexHeatmap")

if (!require("ComplexUpset")) {install.packages("ComplexUpset"); require("ComplexUpset")}

# Install from scratch
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("RSQLite")  # Core dependency
BiocManager::install("org.Mm.eg.db")
BiocManager::install("clusterProfiler")


# CRAN packages
install.packages(c("dplyr", "magrittr"))
library(org.Mm.eg.db)
library(clusterProfiler)
library(AnnotationDbi)
library(dplyr)
library(magrittr)
library(ComplexUpset)
library(ComplexHeatmap)


library("EnhancedVolcano")
library(UpSetR)



```


```{r}
options(future.globals.maxSize = 74 * 1024^3) # 55 GB
getOption("future.globals.maxSize") #59055800320
```

```{r}

load(here("jk_code", "SO4_SUB_analysis.rds"))

SO4 <- SO4[
  !grepl("^(mt-|Rp|Gm)|Rik$", rownames(SO4))
]
```

# Generating DEGs for low_salt vs control

```{r}

# subclass2_MD = 4 types
Idents(SO4) <- "subclass2_MD"

# my List of Subtypes
subtypes <- c('type_1', 'type_2', 'type_3', 'type_4')

# store results
DEGS <- list()

# For loop for find markers in each type
for (subtype in subtypes) {
  df <- FindMarkers(
    object = SO4,
    ident.1 = "low_salt",
    ident.2 = "control",
    group.by = "treatment",
    subset.ident = subtype,
    min.pct = 0.1,
    logfc.threshold = 0.1,
    only.pos = FALSE
  )
  df$gene <- rownames(df)  # Add gene column to the dataframe
  DEGS[[subtype]] <- df  #Store the dataframe in the DEGS list
  
  }
  
# making list to generate upset plot
marker_list <- list(
  type_1 = rownames(DEGS$type_1),
  type_2 = rownames(DEGS$type_2),
  type_3 = rownames(DEGS$type_3),
  type_4 = rownames(DEGS$type_4)
)


treatment_markers <- FindMarkers(
    object = SO4,
    ident.1 = "low_salt",
    ident.2 = "control",
    group.by = "treatment",
    min.pct = 0.1,
    logfc.threshold = 0.1,
    only.pos = FALSE
  )
 treatment_markers$gene <- rownames(treatment_markers)


```

# Grabbing Significant Genes

```{r}

# Generating Unnamed List of Significant Genes
sig_genes <- lapply(subtypes, function(subtype) {
  rownames(DEGS[[subtype]][DEGS[[subtype]]$p_val_adj < 0.05, ])
})

# Naming the genes
names(sig_genes) <- subtypes



```


# Upset Plot

## Upset Plot ( All genes )

```{r}


str(marker_list)

# Create a Matrix
list_to_matrix(marker_list)

G1 <- list_to_matrix(marker_list)

head(G1)

m1 <-  make_comb_mat(marker_list)
m1

#Plot Upset

UpSet(m1, set_order = c("type_1", "type_2","type_3","type_4"), comb_order = order(comb_size(m1))) 


```

## Upset Plot (Significant Genes)

```{r}


str(sig_genes)

# Create a Matrix
list_to_matrix(sig_genes)

G2 <- list_to_matrix(sig_genes)

head(G2)

m2 <-  make_comb_mat(sig_genes)
m2

#Plot Upset

UpSet(m2, set_order = c("type_1", "type_2","type_3","type_4"), comb_order = order(comb_size(m2))) 


```


# Extracting DEGS


## DEGS from (All genes Plot)

```{r}

# m1 = All Genes, m2 = Significant Genes < 0.05

type_1 <- extract_comb(m1,"1000")

type_2 <- extract_comb(m1,"0100")

type_3 <- extract_comb(m1,"0010")

type_4 <- extract_comb(m1,"0001")

type_all <- extract_comb(m1,"1111")


```

# Checking Log change 

Positive avg_log2FC: gene is higher in low_salt than control (upregulated in low_salt)

Negative avg_log2FC: gene is lower in low_salt than control (downregulated in low_salt, or upregulated in control)

```{r}

# Filter for your top genes and arrang, replace top_genes with genes I may want to analyze specificall

# Type 1 
 type1markers <- DEGS$type_1 %>%
  filter(gene %in% type_1) %>%
  arrange(desc(avg_log2FC))

# Type 2
type2markers <- DEGS$type_2 %>%
  filter(gene %in% type_2) %>%
  arrange(desc(avg_log2FC))


# Type 3
type3markers <- DEGS$type_3 %>%
  filter(gene %in% type_3) %>%
  arrange(desc(avg_log2FC))


#Type 4 
type4markers <- DEGS$type_4 %>%
  filter(gene %in% type_4) %>%
  arrange(desc(avg_log2FC))

#IN All
allmarkers <- treatment_markers%>%
  filter(gene %in% type_all) %>%
  arrange(desc(avg_log2FC))


```

# Proportional Plot

```{r}


# Change the code to use the right Seurat Object and column that separates by treatment group.

 Idents(SO4) <- "subclass2_MD"

t1 <- table(Idents(SO4), SO4$treatment)[, c("control", "low_salt")]

t1

 

prop.t1 <- prop.table(t1, margin = 2)

prop.t1

 

t2 <- as.data.frame(t1)

colnames(t2) <- c('MD_subtype', 'Treatment', 'Frequency')

 

# Original plot

 

ggplot(t2, aes(fill=MD_subtype, y=Frequency, x= Treatment)) +

  geom_bar(position="fill", stat = "identity", fun.y = "mean", colour="black") +

  theme_classic()

```

Type 1 -> increases 

Type 2 -> Decreases 

Type 3 -> increase

Type 4 -> increase


