---
title: "Upset Plot Upregulated Analysis"
author: "jeff karnsomprot"
date: "2025-07-15"
output:
  html_document: 
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: yes
    theme: journal
    df_print: paged
    code_folding: hide
    highlight: pygments
---

# Loading in Data sets + Library packages.

```{r load new packages, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require("knitr")) {install.packages("knitr"); require("knitr")}
if (!requireNamespace('BiocManager', quietly = TRUE)) {install.packages('BiocManager'); require("BiocManager")}
if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("stringr")) {install.packages("stringr"); require("stringr")}
if (!require("Seurat")) {install.packages("Seurat"); require("Seurat")}
if (!require("sctransform")) {install.packages("sctransform"); require("sctransform")}
if (!require("glmGamPoi")) {BiocManager::install('glmGamPoi'); require("glmGamPoi")}
if (!require("patchwork")) {install.packages("patchwork"); require("patchwork")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
if (!require("EnhancedVolcano")) {BiocManager::install('EnhancedVolcano'); require("EnhancedVolcano")}
if (!require("DESeq2")) {BiocManager::install('DESeq2'); require("DESeq2")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")}
if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require("RColorBrewer")}
if (!require("car")) {install.packages("car"); require("car")}
if (!require("openxlsx")) {install.packages("openxlsx"); require("openxlsx")}
if (!require("readxl")) {install.packages("readxl"); require("readxl")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("gghighlight")) {install.packages("gghighlight"); require("gghighlight")}
if (!require("ggpmisc")) {install.packages("ggpmisc"); require("ggpmisc")}
if (!require("data.table")) {install.packages("data.table"); require("data.table")}
if (!require("here")) {install.packages("here"); require("here")}
if (!require("clusterProfiler")) {BiocManager::install('clusterProfiler'); require("clusterProfiler")}
if (!require("enrichplot")) {BiocManager::install('enrichplot'); require("enrichplot")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")} # for data frame transformation
if (!require("gplots")) {install.packages("gplots"); require("gplots")}
if (!require("ggvenn")) {install.packages("ggvenn"); require("ggvenn")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("UpSetR")) {install.packages("UpSetR"); require("UpSetR")}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ComplexHeatmap")

if (!require("ComplexUpset")) {install.packages("ComplexUpset"); require("ComplexUpset")}

# Install from scratch
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("RSQLite")  # Core dependency
BiocManager::install("org.Mm.eg.db")
BiocManager::install("clusterProfiler")


# CRAN packages
install.packages(c("dplyr", "magrittr"))
library(org.Mm.eg.db)
library(clusterProfiler)
library(AnnotationDbi)
library(dplyr)
library(magrittr)
library(ComplexUpset)
library(ComplexHeatmap)


library("EnhancedVolcano")
library(UpSetR)



```

```{r}
options(future.globals.maxSize = 74 * 1024^3) # 55 GB
getOption("future.globals.maxSize") #59055800320
```

```{r}

load(here("jk_code", "SO4_SUB_NEW_analysis.rds"))

SO4 <- SO4[
  !grepl("^(mt-|Rp|Gm)|Rik$", rownames(SO4))
]

```

# Generating DEGs for low_salt vs control

```{r}

# subclass2_MD = 4 types
Idents(SO4) <- "subclass2_MD"

# my List of Subtypes
subtypes <- c('type_1', 'type_2', 'type_3', 'type_4')

# store results
DEGS <- list()

# For loop for find markers in each type
for (subtype in subtypes) {
  df <- FindMarkers(
    object = SO4,
    ident.1 = "low_salt",
    ident.2 = "control",
    group.by = "treatment",
    subset.ident = subtype,
    min.pct = 0.1,
    logfc.threshold = 0.1,
    only.pos = TRUE
  )
  df$gene <- rownames(df)  # Add gene column to the dataframe
  DEGS[[subtype]] <- df  #Store the dataframe in the DEGS list
  
  }
  
# making list to generate upset plot
marker_list <- list(
  type_1 = rownames(DEGS$type_1),
  type_2 = rownames(DEGS$type_2),
  type_3 = rownames(DEGS$type_3),
  type_4 = rownames(DEGS$type_4)
)

treatment_markers <- FindMarkers(
    object = SO4,
    ident.1 = "low_salt",
    ident.2 = "control",
    group.by = "treatment",
    min.pct = 0.1,
    logfc.threshold = 0.1,
    only.pos = TRUE
  )
 treatment_markers$gene <- rownames(treatment_markers)


```
# Gyaramati Genes from Paper

```{r}

gyarmati <- c("Fabp3", "Egf", "Ccn1", "Foxq1", "Cxcl12", "Vash2", "Pamr1", "Vegfa", "Ccn3", "Bmp3", "Fgf9", "Spp1", "Wnt10a", "Sfrp1", "Tcf4", "Pappa2", "Unc5d", "Sema3c", "Robo2", "Slit2", "Egfl6", "Hgfac", "Pdgfc", "Megf9", "Frem1", "Thsd4", "Spock2", "Mmp14", "Adamtsl2", "Car8", "Irx1", "Irx2", "Hoxd11", "Hoxc4", "Etv1")


```

# Grabbing Significant Genes

```{r}

# Generating Unnamed List of Significant Genes
sig_genes <- lapply(subtypes, function(subtype) {
  rownames(DEGS[[subtype]][DEGS[[subtype]]$p_val_adj < 1, ])
})

# Naming the genes
names(sig_genes) <- subtypes



```


# Upset Plot 

## Upset Plot ( All genes no Filter P- Val) 

```{r}


str(marker_list)

# Create a Matrix
list_to_matrix(marker_list)

G1 <- list_to_matrix(marker_list)

head(G1)

m1 <-  make_comb_mat(marker_list)
m1

#Plot Upset

UpSet(m1, set_order = c("type_1", "type_2","type_3","type_4"), comb_order = order(comb_size(m1))) 


```

## Upset Plot (Significant Genes P_val < 1)

```{r}


str(sig_genes)

# Create a Matrix
list_to_matrix(sig_genes)

G2 <- list_to_matrix(sig_genes)

head(G2)

m2 <-  make_comb_mat(sig_genes)
m2

#Plot Upset

UpSet(m2, set_order = c("type_1", "type_2","type_3","type_4"), comb_order = order(comb_size(m2))) 


```


# Extracting DEGS


## DEGS from (All genes Plot)

```{r}

# m1 = All Genes, m2 = Significant Genes < 1

type_1 <- extract_comb(m2,"1000")

type_2 <- extract_comb(m2,"0100")

type_3 <- extract_comb(m2,"0010") 

type_4 <- extract_comb(m2,"0001") # type 123

type_12 <- extract_comb(m2,"1100") 

type_all <- extract_comb(m2,"1111")



```

# Checking Log change 

Positive avg_log2FC: gene is higher in low_salt than control (upregulated in low_salt)

Negative avg_log2FC: gene is lower in low_salt than control (downregulated in low_salt, or upregulated in control)


```{r}

# Filter for your top genes and arrang, replace top_genes with genes I may want to analyze specificall

# Type 1 
 type1markers <- DEGS$type_1 %>%
  filter(gene %in% type_1,) %>%
  arrange(desc(avg_log2FC))

# Type 2
type2markers <- DEGS$type_2 %>%
  filter(gene %in% type_2) %>%
  arrange(desc(avg_log2FC))

# Type 3
type3markers <- DEGS$type_3 %>%
  filter(gene %in% type_3) %>%
  arrange(desc(avg_log2FC))

type4markers <- DEGS$type_4 %>%
  filter(gene %in% type_4) %>%
  arrange(desc(avg_log2FC)) 

#IN All
typeallmarkers <- treatment_markers%>%
  filter(gene %in% type_all) %>%
  arrange(desc(avg_log2FC)) 

type12markers <- treatment_markers%>%
  filter(gene %in% type_12) %>%
  arrange(desc(avg_log2FC)) 

```


# Pathway Analysis Function

```{r}

# Optional P-val Filter, Filter Before Independently before you Run this

# Creating a function to Repeat Easily

pathway_analysis <- function(type_markers, type_name)
  {
  # Arrange markers by log2 fold change
  df2 <- type_markers %>% arrange(desc(avg_log2FC))
  DEG_list <- df2
  markers <- DEG_list %>% rownames_to_column(var = "SYMBOL")
  
  
  # Convert SYMBOL to ENTREZID
  ENTREZ_list <- bitr(
    geneID = rownames(DEG_list),
    fromType = "SYMBOL",
    toType = "ENTREZID",
    OrgDb = org.Mm.eg.db
  )
  markers <- ENTREZ_list %>% inner_join(markers, by = "SYMBOL")
  
  # Upregulated Pathway Code
  pos_markers_up <- markers %>% dplyr::filter(avg_log2FC > 0) %>% arrange(desc(abs(avg_log2FC)))
  pos_ranks_up <- pos_markers_up$ENTREZID[abs(pos_markers_up$avg_log2FC) > 0]
  pos_go_up <- enrichGO(gene = pos_ranks_up, OrgDb = "org.Mm.eg.db", ont = "BP", readable = TRUE)
  chart_up <- dotplot(pos_go_up) +
    ggtitle(paste(type_name, "Upregulated Lowsalt")) +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5),
      legend.position = "left",
      axis.text.y = element_text(hjust = 0, size = 10)
    ) +
    scale_y_discrete(position = "right", labels = function(x) str_wrap(x, width = 25))
  

  return(list(
    chart_up = chart_up,
    pos_go_up = pos_go_up
  ))
}
```


# Displaying Results via Pathway Analysis

```{r}


results_type1 <- pathway_analysis(subset(type1markers, avg_log2FC > 0.2), "type 1")
results_type2 <- pathway_analysis(subset(type2markers, avg_log2FC > 0), "type 2")
results_type3 <- pathway_analysis(subset(type3markers, avg_log2FC > 0), "type 3")
results_type4 <- pathway_analysis(subset(type4markers, avg_log2FC > 0), "type 4")
results_typeall <- pathway_analysis(subset(typeallmarkers, avg_log2FC > 0), "type all")
results_type12 <- pathway_analysis(subset(type12markers, avg_log2FC > 0), "type 1 and 2 ")

results_type1
results_type2
results_type3
results_type4
results_typeall
results_type12

```


# GENE VIEW

## TYPE 1

```{r}

pathway_type1up <- strsplit(results_type1$pos_go_up$geneID, "/")

names(pathway_type1up) <- results_type1$pos_go_up$Description 

pathway_type1up



```

## TYPE 2

```{r}

pathway_type2up <- strsplit(results_type2$pos_go_up$geneID, "/")

names(pathway_type2up) <- results_type2$pos_go_up$Description 

pathway_type2up



```

## TYPE 3

```{r}

pathway_type3up <- strsplit(results_type3$pos_go_up$geneID, "/")

names(pathway_type3up) <- results_type3$pos_go_up$Description 

pathway_type3up



```

## TYPE 4

```{r}

pathway_type4up <- strsplit(results_type4$pos_go_up$geneID, "/")

names(pathway_type4up) <- results_type4$pos_go_up$Description 

pathway_type4up



```

## TYPE ALL

```{r}

pathway_typeallup <- strsplit(results_typeall$pos_go_up$geneID, "/")

names(pathway_typeallup) <- results_typeall$pos_go_up$Description 

pathway_typeallup



```


```{r}

VlnPlot(SO4,group.by = "subclass2_MD",split.by = "treatment","Cited1")

FeaturePlot(SO4,"Bmp2",split.by = "subclass2_MD")


```