---
title: "upset_plot_wip"
author: "jeff karnsomprot"
date: "2025-07-08"
output:
  html_document: 
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: yes
    theme: journal
    df_print: paged
    code_folding: hide
    highlight: pygments
---

# Loading in Data sets + Library packages.

```{r load new packages, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require("knitr")) {install.packages("knitr"); require("knitr")}
if (!requireNamespace('BiocManager', quietly = TRUE)) {install.packages('BiocManager'); require("BiocManager")}
if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("stringr")) {install.packages("stringr"); require("stringr")}
if (!require("Seurat")) {install.packages("Seurat"); require("Seurat")}
if (!require("sctransform")) {install.packages("sctransform"); require("sctransform")}
if (!require("glmGamPoi")) {BiocManager::install('glmGamPoi'); require("glmGamPoi")}
if (!require("patchwork")) {install.packages("patchwork"); require("patchwork")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
if (!require("EnhancedVolcano")) {BiocManager::install('EnhancedVolcano'); require("EnhancedVolcano")}
if (!require("DESeq2")) {BiocManager::install('DESeq2'); require("DESeq2")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")}
if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require("RColorBrewer")}
if (!require("car")) {install.packages("car"); require("car")}
if (!require("openxlsx")) {install.packages("openxlsx"); require("openxlsx")}
if (!require("readxl")) {install.packages("readxl"); require("readxl")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("gghighlight")) {install.packages("gghighlight"); require("gghighlight")}
if (!require("ggpmisc")) {install.packages("ggpmisc"); require("ggpmisc")}
if (!require("data.table")) {install.packages("data.table"); require("data.table")}
if (!require("here")) {install.packages("here"); require("here")}
if (!require("clusterProfiler")) {BiocManager::install('clusterProfiler'); require("clusterProfiler")}
if (!require("enrichplot")) {BiocManager::install('enrichplot'); require("enrichplot")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")} # for data frame transformation
if (!require("gplots")) {install.packages("gplots"); require("gplots")}
if (!require("ggvenn")) {install.packages("ggvenn"); require("ggvenn")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("UpSetR")) {install.packages("UpSetR"); require("UpSetR")}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ComplexHeatmap")

if (!require("ComplexUpset")) {install.packages("ComplexUpset"); require("ComplexUpset")}

# Remove corrupted installation
remove.packages(c("RSQLite", "org.Mm.eg.db", "clusterProfiler"))

# Install from scratch
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("RSQLite")  # Core dependency
BiocManager::install("org.Mm.eg.db")
BiocManager::install("clusterProfiler")


# CRAN packages
install.packages(c("dplyr", "magrittr"))
library(org.Mm.eg.db)
library(clusterProfiler)
library(AnnotationDbi)
library(dplyr)
library(magrittr)
library(ComplexUpset)
library(ComplexHeatmap)


library("EnhancedVolcano")
library(UpSetR)



```


```{r}
options(future.globals.maxSize = 74 * 1024^3) # 55 GB
getOption("future.globals.maxSize") #59055800320
```

```{r}

load(here("jk_code", "SO4analysis.rds"))
```

# Generating DEGs for low_salt vs control

```{r}

# subclass2_MD = 4 types
Idents(SO4) <- "subclass2_MD"

# my List of Subtypes
subtypes <- c('type_1', 'type_2', 'type_3', 'type_4')

# store results
marker_list <- list()

# For loop for find markers in each type 
for (subtype in subtypes) {
  df <- FindMarkers(
    object = SO4,
    ident.1 = "low_salt",
    ident.2 = "control",
    group.by = "treatment",
    subset.ident = subtype,
    min.pct = 0.1,
    logfc.threshold = 0.25,
    only.pos = FALSE
  )
  
  # Filter for sig genes adjust as needed
  sig_genes <- rownames(df[df$p_val_adj < 0.05, ]) # rownames extract gene names for use in upset plot
  # store in list under subtype name ( type_1.. etc )
  marker_list[[subtype]] <- sig_genes
}


```


# Generating Upset Plot

```{r}


str(marker_list)

# Create a Matrix
list_to_matrix(marker_list)

G1 <- list_to_matrix(marker_list)

G1 <- G1[!grepl("Rik$", rownames(G1)), ]

head(G1)

m1 <-  make_comb_mat(marker_list)
m1

#Plot Upset

UpSet(m1, set_order = c("type_1", "type_2","type_3","type_4"), comb_order = order(comb_size(m1))) 


```


# Extracting DEGS

```{r}


type_1 <- extract_comb(m1,"1000")

type_2 <- extract_comb(m1,"0100")

type_3 <- extract_comb(m1,"0010")

type_4 <- extract_comb(m1,"0001")






```

# Checking Log change 

```{r}

# Initialize an empty list
DEGS <- list()

for (subtype in subtypes) {
  df <- FindMarkers(
    object = SO4,
    ident.1 = "low_salt",
    ident.2 = "control",
    group.by = "treatment",
    subset.ident = subtype,
    min.pct = 0.1,
    logfc.threshold = 0.25,
    only.pos = FALSE
  )
  # Filter for significant genes
  df2 <- df[df$p_val_adj < 0.05, ]
  # Add a column called gene 
  
  df2$gene <- rownames(df2)
  DEGS[[subtype]] <- df2

}
```

Positive avg_log2FC: gene is higher in low_salt than control (upregulated in low_salt)

Negative avg_log2FC: gene is lower in low_salt than control (downregulated in low_salt, or upregulated in control)

```{r}

# Filter for your top genes and arrang, replace top_genes with genes I may want to analyze specificall

# Type 1 
 type1markers <- DEGS$type_1 %>%
  filter(gene %in% type_1) %>%
  arrange(desc(avg_log2FC))

# Type 2
type2markers <- DEGS$type_2 %>%
  filter(gene %in% type_2) %>%
  arrange(desc(avg_log2FC))


# Type 3
type3markers <- DEGS$type_3 %>%
  filter(gene %in% type_3) %>%
  arrange(desc(avg_log2FC))



#Type 4 
type4markers <- DEGS$type_4 %>%
  filter(gene %in% type_4) %>%
  arrange(desc(avg_log2FC))

```

# analysis of expression between samples (upset genes)

## Type 1 Pathway Upregulated = Low salt 

```{r}

df<- type1markers%>% arrange(desc(avg_log2FC))


df2 <- df %>% filter(p_val_adj < 0.05)


DEG_list <- df2

markers1up <- DEG_list %>% rownames_to_column(var="SYMBOL")


ENTREZ_list <- bitr(
  geneID = rownames(DEG_list),
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Mm.eg.db
)

markers1up <-  ENTREZ_list %>% inner_join(markers1up, by = "SYMBOL")

markers1up <-  markers1up %>% dplyr::filter(p_val_adj < 0.05)
#head(markers, n = 50)

pos.markers1up <-  markers1up %>% dplyr::filter(avg_log2FC > 0.5) %>%  arrange(desc(abs(avg_log2FC)))
#change around avg log value 

pos.ranks1up <- pos.markers1up$ENTREZID[abs(pos.markers1up$avg_log2FC) > 0]
#head(pos.ranks)

pos_go1up <- enrichGO(gene = pos.ranks1up,           #a vector of entrez gene id
                   OrgDb = "org.Mm.eg.db",    
                   ont = "BP",
                   readable = TRUE)              #whether mapping gene ID to gene Name

pos_go1up

 chart1up <- dotplot(pos_go1up) +
    ggtitle("type 1 upregulated lowsalt") +
    theme_classic() + 
    theme(
        plot.title = element_text(hjust = 0.5),
        legend.position = "left",
        axis.text.y = element_text(hjust = 0, size = 10)) +
    scale_y_discrete(position = "right", 
                     labels = function(x) str_wrap(x, width = 25))  # Wrap y-axis labels to 2 lines
 

chart1up

```

## Type 2 Pathway Down regulated = Control

```{r}

df <- type2markers %>% arrange(desc(avg_log2FC))

df2 <- df %>% filter(p_val_adj < 0.05)

DEG_list <- df2

markers2up <- DEG_list %>% rownames_to_column(var="SYMBOL")

ENTREZ_list <- bitr(
  geneID = rownames(DEG_list),
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Mm.eg.db
)

markers2up <- ENTREZ_list %>% inner_join(markers2up, by = "SYMBOL")

markers2up <- markers2up %>% dplyr::filter(p_val_adj < 0.05)
#head(markers2up, n = 50)

pos.markers2up <- markers2up %>% dplyr::filter(avg_log2FC < 0) %>% arrange(desc(abs(avg_log2FC)))
#change around avg log value 

pos.ranks2up <- pos.markers2up$ENTREZID[abs(pos.markers2up$avg_log2FC) > 0]
#head(pos.ranks2up)

pos_go2up <- enrichGO(gene = pos.ranks2up,           #a vector of entrez gene id
                   OrgDb = "org.Mm.eg.db",    
                   ont = "BP",
                   readable = TRUE)              #whether mapping gene ID to gene Name

pos_go2up

chart2down <- dotplot(pos_go2up) +
    ggtitle("type 2 downregulated control") +
    theme_classic() + 
    theme(
        plot.title = element_text(hjust = 0.5),
        legend.position = "left",
        axis.text.y = element_text(hjust = 0, size = 10)) +
    scale_y_discrete(position = "right", 
                     labels = function(x) str_wrap(x, width = 25))  # Wrap y-axis labels to 2 lines

chart2down

```

## Type 3 Pathway Down regulated = Control

```{r}

df <- type3markers %>% arrange(desc(avg_log2FC))

df2 <- df %>% filter(p_val_adj < 0.05)

DEG_list <- df2

markers2up <- DEG_list %>% rownames_to_column(var="SYMBOL")

ENTREZ_list <- bitr(
  geneID = rownames(DEG_list),
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Mm.eg.db
)

markers2up <- ENTREZ_list %>% inner_join(markers2up, by = "SYMBOL")

markers2up <- markers2up %>% dplyr::filter(p_val_adj < 0.05)
#head(markers2up, n = 50)

pos.markers2up <- markers2up %>% dplyr::filter(avg_log2FC < 0) %>% arrange(desc(abs(avg_log2FC)))
#change around avg log value 

pos.ranks2up <- pos.markers2up$ENTREZID[abs(pos.markers2up$avg_log2FC) > 0]
#head(pos.ranks2up)

pos_go2up <- enrichGO(gene = pos.ranks2up,           #a vector of entrez gene id
                   OrgDb = "org.Mm.eg.db",    
                   ont = "BP",
                   readable = TRUE)              #whether mapping gene ID to gene Name

pos_go2up

chart3down <- dotplot(pos_go2up) +
    ggtitle("type 3 downregulated control") +
    theme_classic() + 
    theme(
        plot.title = element_text(hjust = 0.5),
        legend.position = "left",
        axis.text.y = element_text(hjust = 0, size = 10)) +
    scale_y_discrete(position = "right", 
                     labels = function(x) str_wrap(x, width = 25))  # Wrap y-axis labels to 2 lines

chart3down

```

## Type 4  Downregulated = control, negative, supressed of low_salt

```{r}


# Arrange and filter DEGs for type 4
df <- type4markers %>% arrange(desc(avg_log2FC))

df2 <- df %>% filter(p_val_adj < 0.05)

DEG_list <- df2

markers4down <- DEG_list %>% rownames_to_column(var="SYMBOL")

ENTREZ_list <- bitr(
  geneID = rownames(DEG_list),
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Mm.eg.db
)

markers4down <- ENTREZ_list %>% inner_join(markers4down, by = "SYMBOL")

markers4down <- markers4down %>% dplyr::filter(p_val_adj < 0.05)
#head(markers4down, n = 50)

pos.markers4down <- markers4down %>% dplyr::filter(avg_log2FC < 0) %>% arrange(desc(abs(avg_log2FC)))
# Adjust avg_log2FC threshold here if you want a stricter cutoff

pos.ranks4down <- pos.markers4down$ENTREZID[abs(pos.markers4down$avg_log2FC) > 0]
#head(pos.ranks4down)

pos_go4down <- enrichGO(
  gene = pos.ranks4down,           # a vector of entrez gene id
  OrgDb = org.Mm.eg.db,    
  ont = "BP",
  readable = TRUE                  # whether mapping gene ID to gene Name
)

pos_go4down

chart4down <- dotplot(pos_go4down) +
  ggtitle("type 4  Downregulated = control, negative, supressed of low_salt") +
  theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "left",
    axis.text.y = element_text(hjust = 0, size = 10)
  ) +
  scale_y_discrete(
    position = "right", 
    labels = function(x) str_wrap(x, width = 25)
  )  # Wrap y-axis labels to 2 lines

chart4down

```

# Proportional Plot

```{r}


# Change the code to use the right Seurat Object and column that separates by treatment group.

 Idents(SO4) <- "subclass2_MD"

t1 <- table(Idents(SO4), SO4$treatment)[, c("control", "low_salt")]

t1

 

prop.t1 <- prop.table(t1, margin = 2)

prop.t1

 

t2 <- as.data.frame(t1)

colnames(t2) <- c('Cell_type', 'Treatment', 'Frequency')

 

# Original plot

 

ggplot(t2, aes(fill=Cell_type, y=Frequency, x= Treatment)) +

  geom_bar(position="fill", stat = "identity", fun.y = "mean", colour="black") +

  theme_classic()

```
Type 1 -> increases 

Type 2 -> Decreases 

Type 3 -> increase

Type 4 -> increase


