---
title: "MaculaSO5"
author: "jeff karnsomprot"
date: "2025-04-18"
output:
  html_document: 
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: yes
    theme: journal
    df_print: paged
    code_folding: hide
    highlight: pygments

---
# Macula Densa Project

## Objectives 

Three Main Goals of this File <br>
Produce Cleaner looking code. <br> 
Identify the amount of clusters there are <br>
Identify the top genes expressed in each of the clusters <br>


## Problems I need to Fix
Save things as RDS file so I dont have to rerun the whole code

# Loading in Data sets + Library packages.

```{r load new packages, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("Seurat")) {install.packages("Seurat"); require("Seurat")}
if (!require("patchwork")) {install.packages("patchwork"); require("patchwork")}
if (!require("knitr")) {install.packages("knitr"); require("knitr")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
if (!require("BiocManager")) {install.packages("BiocManager"); require("BiocManager")}
if (!require("tibble")) {install.packages("tibble"); require("tibble")}
if (!require("ggpmisc")) {install.packages("ggpmisc"); require("ggpmisc")}
if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require("RColorBrewer")} #color
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("DESeq2")) {BiocManager::install('DESeq2'); require("DESeq2")}
if (!require("here")) {install.packages("here"); require("here")}
if (!require("stringr")) {install.packages("stringr"); require("stringr")}
if (!require("car")) {install.packages("car"); require("car")}
if (!require("openxlsx")) {install.packages("openxlsx"); require("openxlsx")}
if (!require("readxl")) {install.packages("readxl"); require("readxl")}
if (!require("data.table")) {install.packages("data.table"); require("data.table")}
if (!require("ggvenn")) {install.packages("ggvenn"); require("ggvenn")}
if (!require("kableExtra")) {install.packages("kableExtra"); require("kableExtra")} # for color brewer
if (!require("gplots")) {install.packages("gplots"); require("gplots")} # for color brewer
if (!require("clusterProfiler")) {BiocManager::install('clusterProfiler'); require("clusterProfiler")}
if (!require("enrichplot")) {BiocManager::install('enrichplot'); require("enrichplot")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")} # for data frame transformation
if (!require("plotly")) {install.packages("plotly"); require("plotly")}
library("EnhancedVolcano")


```


```{r}
options(future.globals.maxSize = 74 * 1024^3) # 55 GB
getOption("future.globals.maxSize") #59055800320
```

``` {r}

SO5 <- LoadSeuratRds(here("jk_code", "SO5.rds"))

head(SO5@meta.data)


```

# Analyzing the SO5 DATASET
``` {r}

DimPlot(SO5,split.by ="sample")
DimPlot(SO5)


```
Based off this I can see that 

SO1-> control
SO2 -> low_salt 
SO3 -> low_salt 
SO4 -> control

```{r}

SO5 <- FindNeighbors(SO5, dims = 1:30, verbose = F)
SO5 <- FindClusters(SO5, resolution = 0.25)

SO5m <- FindAllMarkers(SO5, only.pos = TRUE)
SO5m %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)

SO5m %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 5) %>%
    ungroup() -> top10
DoHeatmap(SO5, features = top10$gene) + NoLegend()
DimPlot(SO5)
FeaturePlot(SO5,"S100g",split.by = "treatment") #cluster
FeaturePlot(SO5,"Junb",split.by = "treatment") # cluster
FeaturePlot(SO5,"Cxcl10",split.by = "treatment") # cluster
FeaturePlot(SO5,"Pappa2",split.by = "treatment")
DimPlot(SO5,split.by = "treatment")

```

# Identifying Clusters

## Cluster 5
``` {r}

# Cxcl10
FeaturePlot(SO5, features = "Cxcl10", split.by = "treatment")

# Ifit1
FeaturePlot(SO5, features = "Ifit1", split.by = "treatment",order = TRUE)

# Isg15
FeaturePlot(SO5, features = "Isg15", split.by = "treatment",order = TRUE)

# Gbp10
FeaturePlot(SO5, features = "Gbp10", split.by = "treatment")

# Ifi47
FeaturePlot(SO5, features = "Ifi47", split.by = "treatment")

```

## Cluster 4

``` {r}

# Fos
FeaturePlot(SO5, features = "Fos", split.by = "treatment")

# Junb
FeaturePlot(SO5, features = "Junb", split.by = "treatment")

# Egr1
FeaturePlot(SO5, features = "Egr1", split.by = "treatment")

# Fosb
FeaturePlot(SO5, features = "Fosb", split.by = "treatment")

# Zfp36
FeaturePlot(SO5, features = "Zfp36", split.by = "treatment")

```

## Cluster 2

```{r}

# Egf
FeaturePlot(SO5, features = "Egf", split.by = "treatment")

# Krt7
FeaturePlot(SO5, features = "Krt7", split.by = "treatment")

# Fabp3
FeaturePlot(SO5, features = "Fabp3", split.by = "treatment")

# Cldn19
FeaturePlot(SO5, features = "Cldn19", split.by = "treatment")

# Tmem52b
FeaturePlot(SO5, features = "Tmem52b", split.by = "treatment")

```

## Cluster 0 

```{r}

# Mcub
FeaturePlot(SO5, features = "Mcub", split.by = "treatment")

# Aard
FeaturePlot(SO5, features = "Aard", split.by = "treatment")

# Fetub
FeaturePlot(SO5, features = "Fetub", split.by = "treatment")


```
Observation : MCUB seems to be the highly defined gene in low salt

## Cluster 1
```{r}

# S100g
FeaturePlot(SO5, features = "S100g", split.by = "treatment")

```
## Cluster 3

```{r}

# Leng9
FeaturePlot(SO5, features = "Leng9", split.by = "treatment")

```
# Subset Cluster 0, 1, 3

```{r}
# SO6 checking another cluster
SO6<- subset(SO5, idents = c("0","1","3"))

SO6 <- FindNeighbors(SO6, dims = 1:30, verbose = F)
SO6 <- FindClusters(SO6, resolution = 0.1)

DimPlot(SO6,split.by = "sample")
SO6m <- FindAllMarkers(SO6, only.pos = TRUE)
SO6m %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 0.5)

DimPlot(SO6,split.by = "sample")

DimPlot(SO6,split.by = "sample", group.by = "treatment" )
head(SO6@meta.data)

FeaturePlot(SO6,"Pappa2",split.by = "sample")


```

## Clusters

```{r}

# S100g
FeaturePlot(SO6, features = "S100g")

# Aard
FeaturePlot(SO6, features = "Aard")

# Mcub
FeaturePlot(SO6, features = "Mcub")
DimPlot(SO6,group.by = "sample",split.by = "sample")
DimPlot(SO6,group.by = "treatment",split.by = "sample")

```

My guess is that these are the same, As you go from control to low_salt the cells start to express different genes. 
How can I test this? 

I think the next step after I figure out something with these clusters is to figure out what each of these top genes do, the functions, and purpose of them.


Next Steps: 
Take a Few steps back and identify S100g Cluster 

Multidimensional Dotplot (top genes that are expressed)

Possibly Filter out Rp genes for 
maybe mt too

Annot objects

Read Janos' article 

Next less : 
DEG list pathway analysis
volcano plot 

# Backtracking
```{r}

DimPlot(SO5)
# Check S100g so i Have to make this into more clusters.

SO5_new <- RunUMAP(SO5, dims = 1:30, verbose = F)
SO5_new <- FindNeighbors(SO5, dims = 1:30, verbose = F)
SO5_new <- FindClusters(SO5, resolution = .4)

# back tracking so an old one
p <- DimPlot(SO5)
w <- DimPlot(SO5_new)
p+w

# I thought 5 4 2 are its own cluster
# So in this it would be cluster 
# 8=5 , 4=5 , 2=2 , 
# Cluster 6 should be the s100g cluster 

SO5_m <- FindAllMarkers(SO5_new, only.pos = TRUE)
SO5_m %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)

SO5_m %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 5) %>%
    ungroup() -> top10
DoHeatmap(SO5_new, features = top10$gene) + NoLegend()
p+w
```

Unknown from this is now still 0, 1,3, 4,


```{r}

markers.to.plot1 <- c(
  "S100g",    # 
  "Atf3",     # 
  "Egr1",     # 
  "Fos",      # 
  "Jun",      #
  "Junb",     #
  "Pappa2",   #
  "Cxcl10",   # 
  "Cldn19",   # 
  "Krt7",     #
  "Egf",      #
  "Aard",
  "Ptger3",
  "Leng9",
  "Ckb",
  "Mcub",
  "Fabp3",
  "Ccn1",
  "Foxq1",
  "Cxcl12",
  "Vash2",
  "Pamr1",
  "Vegfa"
)


                      
                      
DotPlot(SO5_new,
features = markers.to.plot1,
dot.scale = 8,
dot.min = 0,
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5)+
coord_flip()


# I feel like cluster 8 is a contaminate because the it shares a lot of similar genes with my "assumed clusters"
```

# Subsetting
```{r}

SO7 <- subset(SO5_new, idents = "8", invert = TRUE)

SO7 <- RunUMAP(SO7, dims = 1:30, verbose = F)
SO7 <- FindNeighbors(SO7, dims = 1:30, verbose = F)
SO7 <- FindClusters(SO7, resolution = .4)

DimPlot(SO7)

DotPlot(SO7,
features = markers.to.plot1,
dot.scale = 8,
dot.min = 0,
scale = FALSE,   
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5)+
coord_flip()


VlnPlot(SO7, features = "S100g")
VlnPlot(SO7, features = "Cxcl10")
VlnPlot(SO7, features = "Mcub")
VlnPlot(SO7, features = "Aard")
VlnPlot(SO7, features = "Egf")
VlnPlot(SO7, features = "Ptger3")
VlnPlot(SO7, features = "Leng9")
VlnPlot(SO7, features = "Jun")


```



Cluster 0 = Aard, Mcub, also most in Pappa2 

Cluster 1 = Unknown but Leng9 according to VlnPlot

Cluster 2 = Cldn19, Krt7, Egf, Fabp3, Foxq1, Cxcl12

Cluster 3 = Pappa2 , Leng9 , Vegfa?

Cluster 4 = Ptger3, but VlnPlot makes me question.

Cluster 5 = Junb, Jun, Fos, Egr1, Atf3 

Cluster 6 = S100g

Cluster 7 = Cxcl10 

My assumptions have changed now that I have viewed violin plots. 


I think the main clusters are 5, 6, 7, 2 

I think 0, 1, 3, 4 are its own thing but just undergoing change? 

Specifically I think 1 and 0 are the same and 4 and 3 maybe? 

I think cluster 1 turns into 0 in low salt conditions and cluster 4 turns into 3. 
Both 1 and 3 have Leng9 which is something thats expressed during evolutionary something like that.

According to the Multidimensional dotplot 0 and 1 also look pretty identical minus the expression of a few specific genes but that may jsut be the product of change. 

Not sure how to explain 4 and 3 though.

# Analyzing 

``` {r}


DimPlot(SO7,split.by = "treatment")
DimPlot(SO7,split.by = "sample")


```

SO1 and SO4 = Control
SO2 and SO3 = low_salt

Im guessing, the two Control samples were slightly different. One expressing more in each. They then used induced low salt into each of these two controls types and got SO2 and SO3 but they're different because of control.

I think it might be a good idea to analyze maybe just SO1 and SO2 to see the change of 4 turning into 3 and analyzing SO4 and SO3 separate to see cluster 1 changing to 0.

Okay after reading the research paper i think Im entirely wrong. 

```{r}

markers.to.plot2 <- c(
  # Cluster 0
  "Aard",
  "Mcub",
  "Pappa2",

  # Cluster 1
  "Leng9",

  # Cluster 2
  "Cldn19",
  "Krt7",
  "Egf",
  "Fabp3",
  "Foxq1",
  "Cxcl12",
  "Ckb",

  # Cluster 3
  "Vegfa",

  # Cluster 4
  "Ptger3",

  # Cluster 5
  "Junb",
  "Jun",
  "Fos",
  "Egr1",
  "Atf3",
  "Ccn1",

  # Cluster 6
  "S100g",

  # Cluster 7
  "Cxcl10"
)

DotPlot(
  SO7,
  features = markers.to.plot2,
  dot.scale = 8,
  dot.min = 0,
  scale = FALSE,        
  scale.max = 100,
  scale.min = 0,
  col.min = -2.5,
  col.max = 2.5
) + coord_flip()

```
Cluster 7 = Cxcl10 Type 1
Cluster 6 = S100g Type 2
Cluster 5 = Fos Type 3
Cluster 2 = Egf Type 4
Cluster 0,1,3,4 = same Type 5
```{r}

SO8 <- FindNeighbors(SO7, dims = 1:30, verbose = F)
SO8 <- FindClusters(SO7, resolution = 0.45)
DimPlot(SO8)

SO8 <- subset(SO8, idents = c("0", "1", "3", "4"))


SO8 <- FindNeighbors(SO8, dims = 1:30, verbose = F)
SO8 <- FindClusters(SO8, resolution = 1.2)
DimPlot(SO8)

SO8_m <- FindAllMarkers(SO8, only.pos = TRUE)
SO8_m %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)

SO8_m %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 5) %>%
    ungroup() -> top10_
DoHeatmap(SO8, features = top10_$gene) + NoLegend()
DimPlot(SO8)

DotPlot(SO8,
features = markers.to.plot1,
dot.scale = 8,
dot.min = 0,
scale = FALSE,   
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5)+
coord_flip()

VlnPlot(SO8,"Ptger3")
VlnPlot(SO8,"Ckb")
VlnPlot(SO8,"Aard")
VlnPlot(SO8,"Mcub")
VlnPlot(SO8,"Pappa2")
VlnPlot(SO8,"Leng9")

```

NEXT STEP :

Combine clusters 0, 1, 3, 4

Gyarmati version

my Version

Cluster 7 = Cxcl10 Type 1

Cluster 6 = S100g Type 2

Cluster 5 = Fos Type 3

Cluster 2 = Egf Type 4

Cluster 0,1,3,4 = same Type 5


```{r}
#SO7 Dimplot

SO_7 <- SO7

SO_7@meta.data <- SO_7@meta.data %>% 
  mutate(subclass_MD = dplyr::case_when(
    seurat_clusters == 0  ~ "type_1",
    seurat_clusters == 1  ~ "type_1",
    seurat_clusters == 2  ~ "type_2",
    seurat_clusters == 3  ~ "type_1",
    seurat_clusters == 4  ~ "type_1",
    seurat_clusters == 5  ~ "type_3",
    seurat_clusters == 6  ~ "type_4",
    seurat_clusters == 7  ~ "type_5"
  ))

SO7@meta.data$subclass_MD <- factor(SO_7@meta.data$subclass_MD , levels = c("type_1", "type_2", "type_3", "type_4", "type_5"))

Idents(SO_7) <- SO_7@meta.data$subclass_MD

DimPlot(object = SO_7, reduction = "umap", group.by = "subclass_MD", label = TRUE)


DimPlot(object = SO_7, reduction = "umap", label = TRUE)


markers.to.plot_MD <- c(
  # type_1
  # type_2
  "Egf",
  "Fabp3",
  "Ktr7",

  # type_3
  "Fos",
  "Egr1",
  "Atf3",

  # type_4
  "S100g",

  # type_5
  "Cxcl10"
)

DotPlot(SO_7,
features = markers.to.plot_MD,
dot.scale = 8,
dot.min = 0,
scale = FALSE,
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5)+
coord_flip()

DimPlot(SO_7, split.by = "treatment")

saveRDS(SO_7, file = "SO_7_recent.RDS")

```