---
title: "Rmarkdown Subset Analysis"
author: "jeff karnsomprot"
date: "2025-03-17"
output: html_document
---
```{r load new packages, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("Seurat")) {install.packages("Seurat"); require("Seurat")}
if (!require("patchwork")) {install.packages("patchwork"); require("patchwork")}
if (!require("knitr")) {install.packages("knitr"); require("knitr")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
if (!require("BiocManager")) {install.packages("BiocManager"); require("BiocManager")}
if (!require("tibble")) {install.packages("tibble"); require("tibble")}
if (!require("ggpmisc")) {install.packages("ggpmisc"); require("ggpmisc")}
if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require("RColorBrewer")} #color
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("DESeq2")) {BiocManager::install('DESeq2'); require("DESeq2")}
if (!require("here")) {install.packages("here"); require("here")}
if (!require("stringr")) {install.packages("stringr"); require("stringr")}
if (!require("car")) {install.packages("car"); require("car")}
if (!require("openxlsx")) {install.packages("openxlsx"); require("openxlsx")}
if (!require("readxl")) {install.packages("readxl"); require("readxl")}
if (!require("data.table")) {install.packages("data.table"); require("data.table")}
if (!require("ggvenn")) {install.packages("ggvenn"); require("ggvenn")}
if (!require("kableExtra")) {install.packages("kableExtra"); require("kableExtra")} # for color brewer
if (!require("gplots")) {install.packages("gplots"); require("gplots")} # for color brewer
if (!require("clusterProfiler")) {BiocManager::install('clusterProfiler'); require("clusterProfiler")}
if (!require("enrichplot")) {BiocManager::install('enrichplot'); require("enrichplot")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")} # for data frame transformation
if (!require("plotly")) {install.packages("plotly"); require("plotly")}
library("EnhancedVolcano")


```


```{r}
options(future.globals.maxSize = 74 * 1024^3) # 55 GB
getOption("future.globals.maxSize") #59055800320
```

``` {r}

SO1 <- LoadSeuratRds(here("outputs", "so_merge_raw.rds"))

head(SO1@meta.data)
```


```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.width=10}

SO.markers <- FindAllMarkers(SO1, only.pos = TRUE)
SO.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)

SO.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 5) %>%
    ungroup() -> top10
DoHeatmap(SO1, features = top10$gene) + NoLegend()

markers.to.plot1 <- c("Lrp2",         # PT
                      "Slc5a12",      # PT-S1
                      "Slc13a3",      # PT-S2
                      "Slc16a9",      # PT-S3
                      "Havcr1",       # Injured PT
                      "Epha7",        # dTL
                      "Cryab",        # dTL
                      "Cdh13",        # dTL1
                      "Slc14a2",      # dTL2
                      "Slc12a1",      # TAL
                      "Umod",         # TAL, DCT1
                      "Egf",          # TAL, DCT1,
                      "Cldn10",       # TAL
                      "Cldn16",       # TAL
                      "Nos1",         # MD
                      "Slc12a3",      # DCT
                      "Pvalb",        # DCT1
                      "Slc8a1",       # DCT2, CNT
                      "Aqp2",         # PC
                      "Slc4a1",       # IC-A
                      "Slc26a4",      # IC-B
                      "Nphs1",        # Podo
                      "Ncam1",        # PEC
                      "Flt1",         # Endo
                      "Emcn",         # Glom Endo
                      "Kdr",          # Capillary Endo
                      "Pdgfrb",       # Perivascular
                      "Pdgfra",       # Fib
                      "Piezo2",       # Mesangial
                      "Acta2",        # Mural
                      "Ptprc",        # Immune
                      "Cd74",         # Macrophage
                      "Skap1",        # B/T Cells 
                      "Upk1b",        # Uro
                      "Top2a"         # Proliferation
)
                      
DotPlot(SO1,
features = markers.to.plot1,
dot.scale = 8,
dot.min = 0,
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5)+
coord_flip()

DimPlot(SO1)

```
Macula Densa Cells would Express a lot of Nos1
In this case only clusters 0-3 express a lot of Nos1 and also some TAL which would be expressed a lot at the macula densa. 

Cluster Identification

Cluster 0 - MD
Cluster 1 - MD - TAL 
Cluster 2 - MD 
Cluster 3 - MD
Cluster 4 - Glom Endo
Cluster 5 - Principal Cell
Cluster 6 - IC
Cluster 7 - PT

Making Sure 
```{r}

FeaturePlot(SO1, "Pappa2",split.by = "treatment")
FeaturePlot(SO1, "Nos1",split.by = "treatment")

VlnPlot(SO1, features = "Pappa2",split.by = "sample")

FeaturePlot(SO1, "Aqp2",split.by = "treatment")

FeaturePlot(SO1, "Lrp2",split.by = "treatment")


```

``` {r}

SO2 <- subset(SO1, idents = c("4", "5", "6", "7"), invert = TRUE)

DimPlot(SO2)

SO2 <- SCTransform(SO2, verbose = F)
SO2 <- RunPCA(SO2, verbose = F)
ElbowPlot(SO2)

SO2 <- RunUMAP(SO2, dims = 1:30, verbose = F)
SO2 <- FindNeighbors(SO2, dims = 1:30, verbose = F)
SO2 <- FindClusters(SO2, resolution = .5)

DimPlot(SO2)

SO.markers2 <- FindAllMarkers(SO2, only.pos = TRUE)
SO.markers2 %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)

SO.markers2 %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 5) %>%
    ungroup() -> top10_S2
DoHeatmap(SO2, features = top10_S2$gene) + NoLegend()

markers.to.plot1 <- c("Lrp2",         # PT
                      "Slc5a12",      # PT-S1
                      "Slc13a3",      # PT-S2
                      "Slc16a9",      # PT-S3
                      "Havcr1",       # Injured PT
                      "Epha7",        # dTL
                      "Cryab",        # dTL
                      "Cdh13",        # dTL1
                      "Slc14a2",      # dTL2
                      "Slc12a1",      # TAL
                      "Umod",         # TAL, DCT1
                      "Egf",          # TAL, DCT1,
                      "Cldn10",       # TAL
                      "Cldn16",       # TAL
                      "Nos1",         # MD
                      "Slc12a3",      # DCT
                      "Pvalb",        # DCT1
                      "Slc8a1",       # DCT2, CNT
                      "Aqp2",         # PC
                      "Slc4a1",       # IC-A
                      "Slc26a4",      # IC-B
                      "Nphs1",        # Podo
                      "Ncam1",        # PEC
                      "Flt1",         # Endo
                      "Emcn",         # Glom Endo
                      "Kdr",          # Capillary Endo
                      "Pdgfrb",       # Perivascular
                      "Pdgfra",       # Fib
                      "Piezo2",       # Mesangial
                      "Acta2",        # Mural
                      "Ptprc",        # Immune
                      "Cd74",         # Macrophage
                      "Skap1",        # B/T Cells 
                      "Upk1b",        # Uro
                      "Top2a",
                      "Pappa2",
                      "Ptgs2", # Proliferation
                      "Jun"
)
                      
                      
DotPlot(SO2,
features = markers.to.plot1,
dot.scale = 8,
dot.min = 0,
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5)+
coord_flip()

DimPlot(SO2)



#Nos1
                    #  "Pappa2",       # Macula Densa
                    #  "Ptgs2",        # Macula Densa
                     # "Slc12a1",
                     # "Cxcl14",
                     # "Sema3c",
                     # "Fgf9",
                    #  "Egfl6"

```
According to the Heatmap Cluster 2 and 7 Display similar top genes,
According to the Multidimensional DotPlot Cluster 10 has a good amount of Cryab which is a marker for dTL


PLAN after visiting : deep dive into cluster 10, and maybe clean it further 


``` {r}

FeaturePlot(SO2,"Pappa2",split.by =  "treatment")
FeaturePlot(SO2, "Nos1",split.by = "treatment")
DimPlot(SO2,split.by = "sample",group.by = "sample")

FeaturePlot(SO2,"Ptgs2",split.by =  "treatment")
FeaturePlot(SO2,"Slc12a1",split.by =  "treatment")
FeaturePlot(SO2,"Fgf9",split.by =  "treatment")

FeaturePlot(SO2,"Fosb",split.by =  "treatment")
FeaturePlot(SO2,"Cldn5",split.by =  "treatment")

DimPlot(SO2, group.by = "treatment")
DimPlot(SO2)



```


``` {r}

FeaturePlot(SO2,"Cryab",split.by =  "sample")
FeaturePlot(SO2,"Epha7",split.by =  "sample")


```

``` {r}

SO_10 <- subset(SO2, idents = c("10"))

SO_10 <- RunUMAP(SO_10, dims = 1:30, verbose = F)
SO_10 <- FindNeighbors(SO_10, dims = 1:30, verbose = F)
SO_10 <- FindClusters(SO_10, resolution = .5)

SO.markers2 <- FindAllMarkers(SO_10, only.pos = TRUE)
SO.markers2 %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)

SO.markers2 %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 5) %>%
    ungroup() -> top10_S2
DoHeatmap(SO_10, features = top10_S2$gene) + NoLegend()
 DimPlot(SO_10)               
 
FeaturePlot(SO_10,"Slc12a1",split.by =  "treatment")
FeaturePlot(SO_10,"Cryab",split.by =  "treatment")

DotPlot(SO_10,
features = markers.to.plot1,
dot.scale = 8,
dot.min = 0,
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5)+
coord_flip()

```


So The cells here that express Jun dont really express Slc12a1 and vice versa, deciding to go see connections between Jun and Slc12a1.


Clusters 6, 7, 10 show Cryab a tiny bit





``` {r}

DimPlot(SO2)
FeaturePlot(SO2,"Junb")

library(plotly)

dim_plot <- DimPlot(SO2, reduction = "umap") +
  ggtitle("UMAP Dimensional Reduction")
interactive_plot <- ggplotly(dim_plot)

# Display the interactive plot
interactive_plot

```
I feel like the clusters are 5 

9 is its own cluster
2 and 7 are own cluster
6 is its own cluster
1 and 4 maybe
0 and 3 5



``` {r}


2222
SO3 <- FindNeighbors(SO2, dims = 1:30, verbose = F)
SO3 <- FindClusters(SO2, resolution = 1)


SO.markers3 <- FindAllMarkers(SO3, only.pos = TRUE)
SO.markers3 %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)

SO.markers3 %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 5) %>%
    ungroup() -> top10_S3
DoHeatmap(SO3, features = top10_S3$gene) + NoLegend()
DimPlot(SO3)
dim_plot <- DimPlot(SO3, reduction = "umap") +
  ggtitle("UMAP Dimensional Reduction")
interactive_plot <- ggplotly(dim_plot)

# Display the interactive plot
interactive_plot

FeaturePlot(SO3,"Cxcl10")
FeaturePlot(SO3,"S100g")
# Only in a specific Cluster

```

``` {r}
SO4 <- FindNeighbors(SO2, dims = 1:30, verbose = F)
SO4 <- FindClusters(SO2, resolution = 0.35)

DimPlot(SO4)
DotPlot(SO4,
features = markers.to.plot1,
dot.scale = 8,
dot.min = 0,
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5)+
coord_flip()

SO5 <- subset(SO4, idents = c("0","1","2","3","4","5"))
SO5 <- FindNeighbors(SO5, dims = 1:30, verbose = F)
SO5 <- FindClusters(SO5, resolution = 0.25)
DimPlot(SO5)
DotPlot(SO5,
features = markers.to.plot1,
dot.scale = 8,
dot.min = 0,
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5)+
coord_flip()

SO5m <- FindAllMarkers(SO5, only.pos = TRUE)
SO5m %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)

SO5m %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 5) %>%
    ungroup() -> top10
DoHeatmap(SO5, features = top10$gene) + NoLegend()
DimPlot(SO5)
FeaturePlot(SO5,"S100g",split.by = "treatment") #cluster
FeaturePlot(SO5,"Junb",split.by = "treatment") # cluster
FeaturePlot(SO5,"Cxcl10",split.by = "treatment") # cluster
FeaturePlot(SO5,"Pappa2",split.by = "treatment")
DimPlot(SO5,split.by = "treatment")

```

```{r}
# SO6 checking another cluster
SO6<- subset(SO5, idents = c("0","1","3"))

SO6 <- FindNeighbors(SO6, dims = 1:30, verbose = F)
SO6 <- FindClusters(SO6, resolution = 0.2)

DimPlot(SO6,split.by = "sample")
SO6m <- FindAllMarkers(SO6, only.pos = TRUE)
SO6m %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)

SO6m %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 20) %>%
    ungroup() -> top10_6
DoHeatmap(SO6, features = top10_6$gene) + NoLegend()

```
