---
title: "Processing Pipseq Slc12a1 Lineage MD Cells"
author: "Jonathan Nelson"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document: 
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: yes
    theme: bootstrap
    df_print: paged
    code_folding: hide
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load new packages, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("Seurat")) {install.packages("Seurat"); require("Seurat")}
if (!require("patchwork")) {install.packages("patchwork"); require("patchwork")}
if (!require("knitr")) {install.packages("knitr"); require("knitr")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
if (!require("BiocManager")) {install.packages("BiocManager"); require("BiocManager")}
if (!require("tibble")) {install.packages("tibble"); require("tibble")}
if (!require("ggpmisc")) {install.packages("ggpmisc"); require("ggpmisc")}
if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require("RColorBrewer")} #color
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("DESeq2")) {BiocManager::install('DESeq2'); require("DESeq2")}
if (!require("here")) {install.packages("here"); require("here")}
if (!require("stringr")) {install.packages("stringr"); require("stringr")}
if (!require("car")) {install.packages("car"); require("car")}
if (!require("openxlsx")) {install.packages("openxlsx"); require("openxlsx")}
if (!require("readxl")) {install.packages("readxl"); require("readxl")}
if (!require("data.table")) {install.packages("data.table"); require("data.table")}
if (!require("ggvenn")) {install.packages("ggvenn"); require("ggvenn")}
if (!require("kableExtra")) {install.packages("kableExtra"); require("kableExtra")} # for color brewer
if (!require("gplots")) {install.packages("gplots"); require("gplots")} # for color brewer
if (!require("clusterProfiler")) {BiocManager::install('clusterProfiler'); require("clusterProfiler")}
if (!require("enrichplot")) {BiocManager::install('enrichplot'); require("enrichplot")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")} # for data frame transformation
library("EnhancedVolcano")
library(UpSetR)
library(ComplexHeatmap)
#install.packages("ggtext")
library(ggtext)
library(CellChat)
library(patchwork)
library(sjmisc)
options(stringsAsFactors = FALSE)

here()

```

# Slc12a1-Lineage Pip-seq

## Load Dataset

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE}

SO_TAL <- readRDS(here("datasets", "TAL_PIP.rds"))

head(SO_TAL@meta.data)

Idents(SO_TAL) <- SO_TAL@meta.data$class.TAL

DimPlot(SO_TAL)

```

## Subset MD Cells

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE}


SO_TAL2 <- subset(x = SO_TAL, subset = class.TAL == "MD")

DimPlot(SO_TAL2)

SO_TAL2

DimPlot(SO_TAL2, group.by = "Sample")

```

## Recluster MD Cells

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE}

SO_TAL3 <- SCTransform(SO_TAL2) %>%
    RunPCA() %>%
    FindNeighbors(dims = 1:30) %>%
    FindClusters(resolution = 1) %>%
    RunUMAP(dims = 1:30)

DimPlot(SO_TAL3)

```

## FindMarkers

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.height=8}


df <- FindAllMarkers(SO_TAL3, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

df

df %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > .25) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10

DoHeatmap(SO_TAL3, features = top10$gene) + NoLegend()

```


```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE}

library(DT)
library(dplyr)

clusters <- unique(df$cluster)

for (clus in clusters) {
  df_cluster <- df %>% filter(cluster == clus)
  cat(sprintf("## Cluster %s\n\n", clus))
  datatable(df_cluster, options = list(pageLength = 5, scrollX = TRUE))
}

```


## Save object

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE}

saveRDS(SO_TAL3, here("outputs", "TAL_PIP_MD.rds"))

```


Idents(SO_TAL2) <- SO_TAL2@meta.data$orig.ident

head(SO_TAL2@meta.data)

SO_TAL3 <- subset(x = SO_TAL2, downsample = 2110)

DimPlot(SO_TAL3, group.by = "class.TAL")

FeaturePlot(SO_TAL3, features = c("Slc12a1", "Cldn10", "Cldn16", "Nos1", "Umod"))

VlnPlot(SO_TAL3, "Umod")
VlnPlot(SO_TAL3, "Egf")
VlnPlot(SO_TAL3, "Tmem52b")
VlnPlot(SO_TAL3, "Ckb")

VlnPlot(SO_TAL3, "Car15")
VlnPlot(SO_TAL3, "Pappa2")

FeaturePlot(SO_TAL3, "Nos1")

VlnPlot(SO_TAL3, "nCount_RNA")

list <- c("Scn1a", "Scn2a", "Scn3a", "Scn8a",
"Kcnc1", "Kcnc2", "Kcnc3",
"Kcna1", "Kcna2",
"Hcn1", "Hcn2", "Hcn3", "Hcn4",
"Cacna1g", "Cacna1h",
"Tbx3",
"Npy", "Vip", "Pvalb", "Sst",
"Nsmf", "Ndrg4",
"Bdnf", "Ntrk2",
"Foxp1", "Foxp2",
"Tbr1", "Satb2",
"Gabra1", "Gabrg2", "Gabrb3",
"Gria1", "Grin2b",
"Adra2a", "Drd2", "Htr2a",
"Rgs4")

for (i in list) {
  p1 <- VlnPlot(SO_TAL3, i)
  print(p1)
}


list2 <- c("Nos1",        # Nitric oxide synthase, key macula densa signal
"Ptgs2",       # COX-2, involved in prostaglandin production
"Slc12a1",     # NKCC2 transporter for Na-K-2Cl sensing
"Ptger3",      # PGE2 receptor
"Ptger4",      # Alternative PGE2 receptor
"Calb1",       # Calcium signaling
"Tgfb1",       # TGF signaling involvement
"Cox4i2",      # Macula densa-enriched oxidative stress sensor
"Aqp1",        # Water transport
"Clcnkb",      # Chloride channel
"Ren1",        # Renin, indirect signaling through JGA
"Guca2b",      # Uroguanylin (possible natriuretic peptide signaling)
"Pla2g4a",     # Phospholipase A2, upstream of prostaglandin synthesis
"Npr3",        # Natriuretic peptide clearance receptor
"Atp1a1",      # Na+/K+ ATPase alpha 1
"Klk1",        # Kallikrein, vasoactive peptide
"Adrb1",       # Beta-adrenergic receptor 1
"Grk4",        # G protein–coupled receptor kinase (linked to Na sensing)
"Slc9a3r1",    # NHERF1, involved in ion transporter regulation
"Avpr1a"       # Vasopressin receptor
)

for (i in list2) {
  p1 <- VlnPlot(SO_TAL3, i)
  print(p1)
}

```

# Session Info

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE}

sessionInfo()

```


  