---
title: "macula_densa_jeff"
author: "jeff karnsoprot"
date: "2025-03-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE}

if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("Seurat")) {install.packages("Seurat"); require("Seurat")}
if (!require("patchwork")) {install.packages("patchwork"); require("patchwork")}
if (!require("cowplot")) {install.packages("cowplot"); require("cowplot")}
if (!require("ggpubr")) {install.packages("ggpubr"); require("ggpubr")}
if (!require("plotly")) {install.packages("plotly"); require("plotly")}
if (!require("knitr")) {install.packages("knitr"); require("knitr")}
if (!require("htmlwidgets")) {install.packages("htmlwidgets"); require("htmlwidgets")}
if (!require("here")) {install.packages("here"); require("here")}
if (!require("EnhancedVolcano")) {BiocManager::install('EnhancedVolcano'); require("EnhancedVolcano")} # volcano plot
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")} # for titying up data
if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require("RColorBrewer")} # for color brewer
if (!require("sctransform")) {install.packages("sctransform"); require("sctransform")} # for data normalization
if (!require("glmGamPoi")) {BiocManager::install('glmGamPoi'); require("glmGamPoi")} # for data normalization, sctransform
if (!require("openxlsx")) {install.packages("openxlsx"); require("openxlsx")} # to save .xlsx files
if (!require("SoupX")) {install.packages("SoupX"); require("SoupX")}
if (!require("DoubletFinder")) {BiocManager::install('DoubletFinder'); require("DoubletFinder")}
library(plotly)
library(ggvenn)
library(gplots)

set.seed((12345))
here()
set_here("/Users/jeffkarnsomprot/Desktop/github")


```

#Loading Datasets 
Couldnt get the here to work properly from github directory so I moved the datasets into my folder

```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

options(future.globals.maxSize = 1000 * 1024^2)
S1 <- Read10X_h5(here("Sample-2_S2.h5")) #low_salt
S2 <- Read10X_h5(here("4414_S1.h5")) #control
S3 <- Read10X_h5(here("Sample-3_S3.h5")) # low_salt
S4 <- Read10X_h5(here("Sample-4_S4.h5")) #control

```

Im using my past homeworks as reminders of steps to do. specifically ROC #4 even though im not finished 

#Turn each Dataset into a Seurat Object 

Turn each dataset into a Seurat Object and Give each of the groups a metadata indicating control or low_salt so we can analyze and compare them together

Im using Dr.Nelsons metadata merging
```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}


S1 <- CreateSeuratObject(S1)
S2 <- CreateSeuratObject(S2)
S3 <- CreateSeuratObject(S3)
S4 <- CreateSeuratObject(S4)

#This command adds a metadata column called sample for each one and fills it with its identity "S1-S4"

S1$sample <- "S1"
S2$sample <- "S2"
S3$sample <- "S3"
S4$sample <- "S4"

SO_merge <- merge(S1, c(S2, S3, S4))

SO_merge@meta.data <- SO_merge@meta.data %>% mutate(treatment = dplyr::case_when(
sample == "S1" ~ "low_salt",
sample == "S2" ~ "control",
sample == "S3" ~ "low_salt",
sample == "S4" ~ "control",
))

# Johnathan's code for merging meta-data and naming treatment groups.
head(SO_merge@meta.data)
tail(SO_merge@meta.data)

# Now I can discern each sample and treatment from eachother



# nCount_RNA and nFeature_RNA 

#I want to visualize the datasets to maybe get rid of outliers


```

SCT may reduce batch effects and are good for downstream analyses like clustering and visualization


```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}
SO_merge[["percent.mt"]] <- PercentageFeatureSet(SO_merge, pattern = "^mt-")

VlnPlot(SO_merge, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), group.by = "treatment", ncol = 3)

# code from Homework 2


# i don't think I want to filter out much because I believe my data set was already a clean macula densa with maybe a few that arent.

SO_merge <- SCTransform(SO_merge, verbose = F)
SO_merge <- RunPCA(SO_merge, verbose = F)
ElbowPlot(SO_merge)

SO_merge <- RunUMAP(SO_merge, dims = 1:30, verbose = F)
SO_merge <- FindNeighbors(SO_merge, dims = 1:15, verbose = F)
SO_merge <- FindClusters(SO_merge, resolution = .6)

# ROC#4 used dims 30 
f1 <- DimPlot(SO_merge, reduction = "umap", group.by = "sample")
f2 <- DimPlot(SO_merge, reduction = "umap", group.by = "treatment")
f1 + f2

DimPlot(SO_merge, reduction = "umap")
VlnPlot(SO_merge, "nFeature_RNA")
#cluster 13 seems to be not apart of this system.

dim_plot <- DimPlot(SO_merge, reduction = "umap") +
  ggtitle("UMAP Dimensional Reduction")
interactive_plot <- ggplotly(dim_plot)

# Display the interactive plot
interactive_plot



```
```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

DimPlot(SO_merge, reduction = "umap") + ggtitle("Adjusted Pipeline") + theme(plot.title = element_text(hjust = 0.5))

markers.to.plot1 <- c("Nos1",         # Macula Densa
                      "Pappa2",       # Macula Densa
                      "Ptgs2",        # Macula Densa
                      "Slc12a1",
                      "Cxcl14",
                      "Sema3c",
                      "Fgf9",
                      "Egfl6")      # Macula Densa

DotPlot(SO_merge,
        features = markers.to.plot1,
        dot.scale = 8,
        dot.min = 0,
        scale.max = 100,
        scale.min = 0,
        col.min = -2.5,
        scale = FALSE,
        col.max = 2.5)+
  coord_flip()
```
Elbow Plot = the higher dim I use the more noise I'll pick up essentially



```{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}

FeaturePlot(SO_merge, "Nos1", split.by = "treatment")
FeaturePlot(SO_merge, "Pappa2", split.by = "treatment")

VlnPlot(SO_merge, "Pappa2", group.by = "treatment", split.by = "sample")

FeaturePlot(SO_merge, "Nos1", split.by = "sample")
FeaturePlot(SO_merge, "Pappa2", split.by = "sample")

# S1 = low salt 
# S2 = Control 
# S3 = Low Salt
# S4 = Control 

table(SO_merge$seurat_clusters)

SO_merge <- PrepSCTFindMarkers(SO_merge, assay = "SCT", verbose = TRUE)
top_genes_cluster9 <- FindMarkers(SO_merge, 
                                  ident.1 = "9",  # Replace "0" with the cluster of interest
                                  only.pos = TRUE,  # Only return upregulated genes
                                  min.pct = 0.6,   # Minimum % of cells expressing the gene
                                  logfc.threshold = 0.25)  # Minimum log-fold change

top_genes_cluster9 %>% arrange(desc(avg_log2FC)) %>% head(n = 10)

FeaturePlot(SO_merge, features = c("Sox18", "Kcne3", "Palmd"))
VlnPlot(SO_merge, features = c("Sox18", "Kcne3", "Palmd"))

FeaturePlot(SO_merge, features = c("Fabp3", "Vegfa", "Egf"))

```

{r, echo=TRUE, warning=FALSE, error=FALSE, message=FALSE, fig.align = 'center'}


cluster_markers <- FindAllMarkers(SO_merge, 
                                  only.pos = TRUE,       # Only positive markers
                                  min.pct = 0.25,        # Minimum % of cells expressing the gene
                                  logfc.threshold = 0.25) # Minimum log-fold change
top_genes<- cluster_markers %>%
  group_by(cluster) %>%
  top_n(n = 3, wt = avg_log2FC)


``` 

