---
title: "control vs SO4"
author: "jeff karnsomprot"
date: "2025-06-05"
output:
  html_document: 
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: yes
    theme: journal
    df_print: paged
    code_folding: hide
    highlight: pygments
---

# Loading in Packages + Dataset

```{r load new packages, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require("knitr")) {install.packages("knitr"); require("knitr")}
if (!requireNamespace('BiocManager', quietly = TRUE)) {install.packages('BiocManager'); require("BiocManager")}
if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("stringr")) {install.packages("stringr"); require("stringr")}
if (!require("Seurat")) {install.packages("Seurat"); require("Seurat")}
if (!require("sctransform")) {install.packages("sctransform"); require("sctransform")}
if (!require("glmGamPoi")) {BiocManager::install('glmGamPoi'); require("glmGamPoi")}
if (!require("patchwork")) {install.packages("patchwork"); require("patchwork")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
if (!require("EnhancedVolcano")) {BiocManager::install('EnhancedVolcano'); require("EnhancedVolcano")}
if (!require("DESeq2")) {BiocManager::install('DESeq2'); require("DESeq2")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")}
if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require("RColorBrewer")}
if (!require("car")) {install.packages("car"); require("car")}
if (!require("openxlsx")) {install.packages("openxlsx"); require("openxlsx")}
if (!require("readxl")) {install.packages("readxl"); require("readxl")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("gghighlight")) {install.packages("gghighlight"); require("gghighlight")}
if (!require("ggpmisc")) {install.packages("ggpmisc"); require("ggpmisc")}
if (!require("data.table")) {install.packages("data.table"); require("data.table")}
if (!require("here")) {install.packages("here"); require("here")}
if (!require("clusterProfiler")) {BiocManager::install('clusterProfiler'); require("clusterProfiler")}
if (!require("enrichplot")) {BiocManager::install('enrichplot'); require("enrichplot")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")} # for data frame transformation
if (!require("gplots")) {install.packages("gplots"); require("gplots")}
if (!require("ggvenn")) {install.packages("ggvenn"); require("ggvenn")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}


# Install from scratch
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("RSQLite")  # Core dependency
BiocManager::install("org.Mm.eg.db")
BiocManager::install("clusterProfiler")


# CRAN packages
install.packages(c("dplyr", "magrittr"))
library(org.Mm.eg.db)
library(clusterProfiler)
library(AnnotationDbi)
library(dplyr)
library(magrittr)




library("EnhancedVolcano")
library(UpSetR)

```

```{r}

load(here("jk_code", "SO4analysis.rds"))

```

# Isolating control and SO4 samples

```{r}

DimPlot(SO4,split.by = "treatment",group.by = "subclass_MD" )

Idents <- "treatment"
DimPlot(SO4)

Idents(SO4) <- "sample"

DimPlot(SO4,split.by = "sample")

control <- subset(SO4, idents = c("SO1","SO4"))

low_salt <- subset(SO4, idents = c("SO2","SO3"))

DimPlot(SO4)


```

# Finding Markers that define each Control Sample

```{r}

control <- SCTransform(control) %>%
    RunPCA() %>%
    FindNeighbors(dims = 1:30) %>%
    FindClusters() %>%
    RunUMAP(dims = 1:30)

DimPlot(control)
DimPlot(control,split.by = "sample")

Idents(control) <- "sample"
control.markers <- FindAllMarkers(control, only.pos = TRUE)
control.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)

control.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10
DoHeatmap(control, features = top10$gene) + NoLegend()

```

# Finding Markers that define each low salt  Sample

```{r}

low_salt <- SCTransform(low_salt) %>%
    RunPCA() %>%
    FindNeighbors(dims = 1:30) %>%
    FindClusters() %>%
    RunUMAP(dims = 1:30)

DimPlot(low_salt)
DimPlot(low_salt,split.by = "sample")

Idents(low_salt) <- "sample"
low_salt.markers <- FindAllMarkers(control, only.pos = TRUE)
low_salt.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)

low_salt.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10
DoHeatmap(low_salt, features = top10$gene) + NoLegend()

```

# Visualizing Difference

```{r}


FeaturePlot(control,"Ptger3")
FeaturePlot(control, "S100g")

VlnPlot(control,"Hspa8")
VlnPlot(control, "Ptger3")
VlnPlot(control, "S100g")

markers.to.plot1 <- c("Ptger3","Hspa8","S100g")
  

DotPlot(control,
features = markers.to.plot1,
dot.scale = 8,
dot.min = 0,
scale = FALSE,
scale.max = 100,
scale.min = 0,
col.min = -2.5,
col.max = 2.5)+
coord_flip()


```



control expresses highly in Ptger3, SO4 expresses high in S100g.

S01 Also has a lot more Ribosomal gene influence. 

Why? 

# Separating the markers

```{r}

control_markers <- control.markers[control.markers$cluster == "control", ]
SO4_markers <- control.markers[control.markers$cluster == "SO4", ]


```

# View pathway analysis for the genes

## control, Control sample 1
```{r}


df<- control.markers %>% arrange(desc(avg_log2FC))

df2 <- df %>% filter(p_val_adj < 0.05)

DEG_list <- df2

markers <- DEG_list %>% rownames_to_column(var="SYMBOL")

head(markers, n = 50)

ENTREZ_list <- bitr(
  geneID = rownames(DEG_list),
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Mm.eg.db
)


markers <-  ENTREZ_list %>% inner_join(markers, by = "SYMBOL")

# Removing genes that are not statistically significant. 
markers <-  markers %>% dplyr::filter(p_val_adj < 0.05)
#head(markers, n = 50)

pos.markers <-  markers %>% dplyr::filter(avg_log2FC > 0) %>%  arrange(desc(abs(avg_log2FC)))
#head(pos.markers, n = 50)

pos.ranks <- pos.markers$ENTREZID[abs(pos.markers$avg_log2FC) > 0]
#head(pos.ranks)

pos_go <- enrichGO(gene = pos.ranks,           #a vector of entrez gene id
                   OrgDb = "org.Mm.eg.db",    
                   ont = "BP",
                   readable = TRUE)              #whether mapping gene ID to gene Name

pos_go

 dotplot(pos_go) +
    ggtitle("") +
    theme_classic() + 
    theme(
        plot.title = element_text(hjust = 0.5),
        legend.position = "left",
        axis.text.y = element_text(hjust = 0, size = 10)) +
    scale_y_discrete(position = "right", 
                     labels = function(x) str_wrap(x, width = 25))  # Wrap y-axis labels to 2 lines
```
