---
title: "low salt vs Control"
author: "jeff karnsomprot"
date: "2025-07-10"
output:
  html_document: 
    toc: yes
    toc_depth: 3
    toc_float: true
    number_sections: yes
    theme: journal
    df_print: paged
    code_folding: hide
    highlight: pygments
---

# Loading in Data sets + Library packages.

```{r load new packages, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}

options(repos = c(CRAN = "https://cloud.r-project.org"))
if (!require("knitr")) {install.packages("knitr"); require("knitr")}
if (!requireNamespace('BiocManager', quietly = TRUE)) {install.packages('BiocManager'); require("BiocManager")}
if (!require("dplyr")) {install.packages("dplyr"); require("dplyr")}
if (!require("stringr")) {install.packages("stringr"); require("stringr")}
if (!require("Seurat")) {install.packages("Seurat"); require("Seurat")}
if (!require("sctransform")) {install.packages("sctransform"); require("sctransform")}
if (!require("glmGamPoi")) {BiocManager::install('glmGamPoi'); require("glmGamPoi")}
if (!require("patchwork")) {install.packages("patchwork"); require("patchwork")}
if (!require("ggplot2")) {install.packages("ggplot2"); require("ggplot2")}
if (!require("EnhancedVolcano")) {BiocManager::install('EnhancedVolcano'); require("EnhancedVolcano")}
if (!require("DESeq2")) {BiocManager::install('DESeq2'); require("DESeq2")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")}
if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require("RColorBrewer")}
if (!require("car")) {install.packages("car"); require("car")}
if (!require("openxlsx")) {install.packages("openxlsx"); require("openxlsx")}
if (!require("readxl")) {install.packages("readxl"); require("readxl")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}
if (!require("gghighlight")) {install.packages("gghighlight"); require("gghighlight")}
if (!require("ggpmisc")) {install.packages("ggpmisc"); require("ggpmisc")}
if (!require("data.table")) {install.packages("data.table"); require("data.table")}
if (!require("here")) {install.packages("here"); require("here")}
if (!require("clusterProfiler")) {BiocManager::install('clusterProfiler'); require("clusterProfiler")}
if (!require("enrichplot")) {BiocManager::install('enrichplot'); require("enrichplot")}
if (!require("tidyverse")) {install.packages("tidyverse"); require("tidyverse")} # for data frame transformation
if (!require("gplots")) {install.packages("gplots"); require("gplots")}
if (!require("ggvenn")) {install.packages("ggvenn"); require("ggvenn")}
if (!require("ggrepel")) {install.packages("ggrepel"); require("ggrepel")}

# Install from scratch
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("RSQLite")  # Core dependency
BiocManager::install("org.Mm.eg.db")
BiocManager::install("clusterProfiler")


# CRAN packages
install.packages(c("dplyr", "magrittr"))
library(org.Mm.eg.db)
library(clusterProfiler)
library(AnnotationDbi)
library(dplyr)
library(magrittr)




library("EnhancedVolcano")
library(UpSetR)

```


# Loading Dataset

```{r}
options(future.globals.maxSize = 74 * 1024^3) # 55 GB
getOption("future.globals.maxSize") #59055800320
```

```{r}

load(here("jk_code", "SO4analysis.rds"))
```


```{r}

SO4$sample <- factor(SO4$sample, levels = c("SO1", "SO4", "SO3", "SO2"))

VlnPlot(SO4,"S100g",split.by = "treatment",group.by = "sample")

# Macula Densa Markers , Nos1, Avpr1a, Slc12a1, Ptgs2, and Slc9a2

VlnPlot(SO4, c("Nos1","Avpr1a","Slc12a1","Ptgs2","Slc9a2","Pappa2"),split.by = "treatment",group.by = "treatment")

DimPlot(SO4)

SO4 <- SO4[
  !grepl("^(mt-|Rp|Gm)|Rik$", rownames(SO4)),
  ,
  drop = FALSE
]

DimPlot(SO4,group.by = "treatment",split.by = "treatment")

#compare each type 1 control to type 1 low_treatment

```

# Markers of control vs low_salt up/down regulation
 
 Upregulated = low salt, positive, activated by low_salt
 Downregulated = control, negative, supressed of low_salt
 
 
```{r}
Idents(SO4) <- "subclass2_MD"
subtypes <- c('type_1', 'type_2','type_3','type_4')
markers_list <- list()

# forloop markers for each subtype

for (subtype in subtypes) {
  markers_list[[subtype]] <- FindMarkers(
    object = SO4,
    ident.1 = "low_salt",
    ident.2 = "control",
    group.by = "treatment",
    subset.ident = subtype,
    min.pct = 0.1,
    logfc.threshold = 0.25,
    only.pos = FALSE
  )
}



VlnPlot(SO4, features = "Pappa2", group.by = "treatment", split.by = "subclass_MD")


DimPlot(SO4, reduction = "umap", split.by = "treatment", label = TRUE)

table(SO4$subclass2_MD, SO4$treatment)


```


## Grabbing each markers of each type low salt vs control
```{r}

type_1 <- markers_list$type_1
type_2<- markers_list$type_2
type_3 <- markers_list$type_3
type_4 <- markers_list$type_4

```

# Type 1

## Type 1 Pathway Upregulated = Low salt 

```{r}

df<- type_1%>% arrange(desc(avg_log2FC))


df2 <- df %>% filter(p_val_adj < 0.05)


DEG_list <- df2

markers1up <- DEG_list %>% rownames_to_column(var="SYMBOL")


ENTREZ_list <- bitr(
  geneID = rownames(DEG_list),
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Mm.eg.db
)

markers1up <-  ENTREZ_list %>% inner_join(markers1up, by = "SYMBOL")

markers1up <-  markers1up %>% dplyr::filter(p_val_adj < 0.05)
#head(markers, n = 50)

pos.markers1up <-  markers1up %>% dplyr::filter(avg_log2FC > 0.5) %>%  arrange(desc(abs(avg_log2FC)))
#change around avg log value 

pos.ranks1up <- pos.markers1up$ENTREZID[abs(pos.markers1 up$avg_log2FC) > 0]
#head(pos.ranks)

pos_go1up <- enrichGO(gene = pos.ranks1up,           #a vector of entrez gene id
                   OrgDb = "org.Mm.eg.db",    
                   ont = "BP",
                   readable = TRUE)              #whether mapping gene ID to gene Name

pos_go1up

 chart1up <- dotplot(pos_go1up) +
    ggtitle("type 1 upregulated lowsalt") +
    theme_classic() + 
    theme(
        plot.title = element_text(hjust = 0.5),
        legend.position = "left",
        axis.text.y = element_text(hjust = 0, size = 10)) +
    scale_y_discrete(position = "right", 
                     labels = function(x) str_wrap(x, width = 25))  # Wrap y-axis labels to 2 lines
 

chart1up
```


### Type 1 Up Gene View 

```{r}
pathway_type1up <- strsplit(pos_go1up$geneID, "/")

names(pathway_type1up) <- pos_go1up$Description 

pathway_type1up

```

## Type 1 Pathway Downregulated = Control ( absence of low salt) 

```{r}

# Arrange and filter DEGs

df <- type_1 %>% arrange(desc(avg_log2FC))

df2 <- df %>% filter(p_val_adj < 0.05)

DEG_list <- df2

markers1down <- DEG_list %>% rownames_to_column(var="SYMBOL")

ENTREZ_list <- bitr(
  geneID = rownames(DEG_list),
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Mm.eg.db
)

markers1down <- ENTREZ_list %>% inner_join(markers1down, by = "SYMBOL")

markers1down <- markers1down %>% dplyr::filter(p_val_adj < 0.05)

pos.markers1down <- markers1down %>% dplyr::filter(avg_log2FC < 0) %>% arrange(desc(abs(avg_log2FC)))

pos.ranks1down <- pos.markers1down$ENTREZID[abs(pos.markers1down$avg_log2FC) > 0.2]

# GO enrichment for downregulated genes
pos_go1down <- enrichGO(
  gene = pos.ranks1down,
  OrgDb = org.Mm.eg.db,
  ont = "BP",
  readable = TRUE
)

pos_go1down

# Visualization
chart1down <- dotplot(pos_go1down) +
  ggtitle("type 1 downregulated control") +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "left",
    axis.text.y = element_text(hjust = 0, size = 10)
  ) +
  scale_y_discrete(
    position = "right",
    labels = function(x) str_wrap(x, width = 25)
  )

chart1down

```

### Type 1 Down Gene View 

```{r}
pathway_type1down <- strsplit(pos_go1down$geneID, "/")

names(pathway_type1down) <- pos_go1down$Description 

pathway_type1down

```

# Type 2 

## Type 2 Pathway Upregulated = Low salt 

```{r}

df <- type_2 %>% arrange(desc(avg_log2FC))

df2 <- df %>% filter(p_val_adj < 0.05)

DEG_list <- df2

markers2up <- DEG_list %>% rownames_to_column(var="SYMBOL")

ENTREZ_list <- bitr(
  geneID = rownames(DEG_list),
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Mm.eg.db
)

markers2up <- ENTREZ_list %>% inner_join(markers2up, by = "SYMBOL")

markers2up <- markers2up %>% dplyr::filter(p_val_adj < 0.05)
#head(markers2up, n = 50)

pos.markers2up <- markers2up %>% dplyr::filter(avg_log2FC > 0) %>% arrange(desc(abs(avg_log2FC))) %>% head(200)
#change around avg log value 

pos.ranks2up <- pos.markers2up$ENTREZID[abs(pos.markers2up$avg_log2FC) > 0]
#head(pos.ranks2up)

pos_go2up <- enrichGO(gene = pos.ranks2up,           #a vector of entrez gene id
                   OrgDb = "org.Mm.eg.db",    
                   ont = "BP",
                   readable = TRUE)              #whether mapping gene ID to gene Name

pos_go2up

chart2up <- dotplot(pos_go2up) +
    ggtitle("type 2 upregulated lowsalt") +
    theme_classic() + 
    theme(
        plot.title = element_text(hjust = 0.5),
        legend.position = "left",
        axis.text.y = element_text(hjust = 0, size = 10)) +
    scale_y_discrete(position = "right", 
                     labels = function(x) str_wrap(x, width = 25))  # Wrap y-axis labels to 2 lines

chart2up

```
### Gene View 

```{r}
pathway_type2up <- strsplit(pos_go2up$geneID, "/")

names(pathway_type2up) <- pos_go2up$Description 

pathway_type2up

```

## Type 2 Pathway Downregulated = Control ( absence of low salt) 


```{r}


# Arrange and filter DEGs for type 2
df <- type_2 %>% arrange(desc(avg_log2FC))

df2 <- df %>% filter(p_val_adj < 0.05)

DEG_list <- df2

markers2down <- DEG_list %>% rownames_to_column(var="SYMBOL")

ENTREZ_list <- bitr(
  geneID = rownames(DEG_list),
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Mm.eg.db
)

markers2down <- ENTREZ_list %>% inner_join(markers2down, by = "SYMBOL")

markers2down <- markers2down %>% dplyr::filter(p_val_adj < 0.05)

pos.markers2down <- markers2down %>% dplyr::filter(avg_log2FC < 0) %>% arrange(desc(abs(avg_log2FC)))

pos.ranks2down <- pos.markers2down$ENTREZID[abs(pos.markers2down$avg_log2FC) > 0]

# GO enrichment for downregulated genes
pos_go2down <- enrichGO(
  gene = pos.ranks2down,
  OrgDb = org.Mm.eg.db,
  ont = "BP",
  readable = TRUE
)

pos_go2down

# Visualization
chart2down <- dotplot(pos_go2down) +
  ggtitle("type 2 downregulated control") +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "left",
    axis.text.y = element_text(hjust = 0, size = 10)
  ) +
  scale_y_discrete(
    position = "right",
    labels = function(x) str_wrap(x, width = 25)
  )

chart2down

```

### Gene View 

```{r}
pathway_type2down <- strsplit(pos_go2down$geneID, "/")

names(pathway_type2down) <- pos_go2down$Description 

pathway_type2down


```


# Type 3 Upregulated -> lowsalt

```{r}

# Arrange and filter DEGs for type 3
df <- type_3 %>% arrange(desc(avg_log2FC))

df2 <- df %>% filter(p_val_adj < 0.05)

DEG_list <- df2

markers3up <- DEG_list %>% rownames_to_column(var="SYMBOL")

ENTREZ_list <- bitr(
  geneID = rownames(DEG_list),
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Mm.eg.db
)

markers3up <- ENTREZ_list %>% inner_join(markers3up, by = "SYMBOL")

markers3up <- markers3up %>% dplyr::filter(p_val_adj < 0.05)
#head(markers3up, n = 50)

pos.markers3up <- markers3up %>% dplyr::filter(avg_log2FC > 0) %>% arrange(desc(abs(avg_log2FC)))
#change around avg log value 

pos.ranks3up <- pos.markers3up$ENTREZID[abs(pos.markers3up$avg_log2FC) > 0]
#head(pos.ranks3up)

pos_go3up <- enrichGO(
  gene = pos.ranks3up,           # a vector of entrez gene id
  OrgDb = org.Mm.eg.db,    
  ont = "BP",
  readable = TRUE                # whether mapping gene ID to gene Name
)

pos_go3up

chart3up <- dotplot(pos_go3up) +
  ggtitle("type 3 upregulated lowsalt") +
  theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "left",
    axis.text.y = element_text(hjust = 0, size = 10)
  ) +
  scale_y_discrete(
    position = "right", 
    labels = function(x) str_wrap(x, width = 25)
  )  # Wrap y-axis labels to 2 lines

chart3up

```

## Gene View Upregulated Type 3

```{r}

pathway_type3up <- strsplit(pos_go3up$geneID, "/")

names(pathway_type3up) <- pos_go3up$Description 

pathway_type3up

```

## Type 3 Downregulated Control asbence of low salt

```{r}
# Arrange and filter DEGs for type 3
df <- type_3 %>% arrange(desc(avg_log2FC))

df2 <- df %>% filter(p_val_adj < 0.05)

DEG_list <- df

markers3down <- DEG_list %>% rownames_to_column(var="SYMBOL")

ENTREZ_list <- bitr(
  geneID = rownames(DEG_list),
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Mm.eg.db
)

markers3down <- ENTREZ_list %>% inner_join(markers3down, by = "SYMBOL")

markers3down <- markers3down %>% dplyr::filter(p_val_adj < 0.05)
#head(markers3down, n = 50)

pos.markers3down <- markers3down %>% dplyr::filter(avg_log2FC < 0) %>% arrange(desc(abs(avg_log2FC)))
# You can adjust the log2FC threshold here if you want a stricter cutoff

pos.ranks3down <- pos.markers3down$ENTREZID[abs(pos.markers3down$avg_log2FC) > 0]
#head(pos.ranks3down)

pos_go3down <- enrichGO(
  gene = pos.ranks3down,           # a vector of entrez gene id
  OrgDb = org.Mm.eg.db,    
  ont = "BP",
  readable = TRUE                  # whether mapping gene ID to gene Name
)

pos_go3down

chart3down <- dotplot(pos_go3down) +
  ggtitle("type 3 downregulated lowsalt") +
  theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "left",
    axis.text.y = element_text(hjust = 0, size = 10)
  ) +
  scale_y_discrete(
    position = "right", 
    labels = function(x) str_wrap(x, width = 25)
  )  # Wrap y-axis labels to 2 lines

chart3down

```

## Gene View Down regulated 

```{r}


pathway_type3down <- strsplit(pos_go3down$geneID, "/")

names(pathway_type3down) <- pos_go3down$Description 

pathway_type3down

```

# Type 4 

## Type 4 Upregulated lowsalt 

```{r}
# Arrange and filter DEGs for type 4
df <- type_4 %>% arrange(desc(avg_log2FC))

df2 <- df %>% filter(p_val_adj < 0.05)

DEG_list <- df2

markers4up <- DEG_list %>% rownames_to_column(var="SYMBOL")

ENTREZ_list <- bitr(
  geneID = rownames(DEG_list),
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Mm.eg.db
)

markers4up <- ENTREZ_list %>% inner_join(markers4up, by = "SYMBOL")

markers4up <- markers4up %>% dplyr::filter(p_val_adj < 0.05)
#head(markers4up, n = 50)

pos.markers4up <- markers4up %>% dplyr::filter(avg_log2FC > 0) %>% arrange(desc(abs(avg_log2FC)))
# Adjust avg_log2FC threshold as needed

pos.ranks4up <- pos.markers4up$ENTREZID[abs(pos.markers4up$avg_log2FC) > 0]
#head(pos.ranks4up)

pos_go4up <- enrichGO(
  gene = pos.ranks4up,           # a vector of entrez gene id
  OrgDb = org.Mm.eg.db,    
  ont = "BP",
  readable = TRUE                # whether mapping gene ID to gene Name
)

pos_go4up

chart4up <- dotplot(pos_go4up) +
  ggtitle("type 4 upregulated lowsalt") +
  theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "left",
    axis.text.y = element_text(hjust = 0, size = 10)
  ) +
  scale_y_discrete(
    position = "right", 
    labels = function(x) str_wrap(x, width = 25)
  )  # Wrap y-axis labels to 2 lines

chart4up

```

## Gene View Upregulated 

```{r}


pathway_type4up <- strsplit(pos_go4up$geneID, "/")

names(pathway_type4up) <- pos_go4up$Description 

pathway_type4up

```

## Type 4 Downregulated lowsalt 

```{r}


# Arrange and filter DEGs for type 4
df <- type_4 %>% arrange(desc(avg_log2FC))

df2 <- df %>% filter(p_val_adj < 0.05)

DEG_list <- df2

markers4down <- DEG_list %>% rownames_to_column(var="SYMBOL")

ENTREZ_list <- bitr(
  geneID = rownames(DEG_list),
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Mm.eg.db
)

markers4down <- ENTREZ_list %>% inner_join(markers4down, by = "SYMBOL")

markers4down <- markers4down %>% dplyr::filter(p_val_adj < 0.05)
#head(markers4down, n = 50)

pos.markers4down <- markers4down %>% dplyr::filter(avg_log2FC < 0) %>% arrange(desc(abs(avg_log2FC)))
# Adjust avg_log2FC threshold here if you want a stricter cutoff

pos.ranks4down <- pos.markers4down$ENTREZID[abs(pos.markers4down$avg_log2FC) > 0]
#head(pos.ranks4down)

pos_go4down <- enrichGO(
  gene = pos.ranks4down,           # a vector of entrez gene id
  OrgDb = org.Mm.eg.db,    
  ont = "BP",
  readable = TRUE                  # whether mapping gene ID to gene Name
)

pos_go4down

chart4down <- dotplot(pos_go4down) +
  ggtitle("type 4 downregulated lowsalt") +
  theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "left",
    axis.text.y = element_text(hjust = 0, size = 10)
  ) +
  scale_y_discrete(
    position = "right", 
    labels = function(x) str_wrap(x, width = 25)
  )  # Wrap y-axis labels to 2 lines

chart4down

```


